---
title: "Analysis of case category status"
author: "UWCHR"
date: "2025-07-02"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, ggforce, plotly, gghighlight, viridis, readxl, ggVennDiagram)

```

Note arrests includes only ICE arrests.

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Admin_Arrests.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names()

arrest_df <- arrest_df %>% 
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("unique_identifier")) |>
  filter(!is.na(unique_identifier))

arrest_ids <- unique(arrest_df$unique_identifier)

latest_arrest <- arrest_df %>% 
  arrange(desc(apprehension_date)) %>% 
  distinct(unique_identifier, .keep_all = TRUE)

```

Note removals includes ICE and CBP apprehensions, shorter timeline.

```{r setup_removals, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ICE_Removals_2025.xlsx"

removals_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(n_dupe = row_number(), .by = c("unique_identifier")) %>% 
  filter(!is.na(unique_identifier))

removal_ids <- unique(removals_df$unique_identifier)

latest_removal <- removals_df %>% 
  arrange(desc(departed_date)) %>% 
  distinct(unique_identifier, .keep_all = TRUE)

```

```{r category_factor_order}

category_order <- c("[1A] Voluntary Departure - Un-Expired and Un-Extended Departure Period",
                    "[1B] Voluntary Departure - Extended Departure Period",
                    "[2A] Deportable - Under Adjudication by IJ",
                    "[2B] Deportable - Under Adjudication by BIA",
                    "[3] Deportable - Administratively Final Order",
                    "[5A] Referred for Investigation - No Show for Hearing - No Final Order",
                    "[5B] Removable - ICE Fugitive",
                    "[5C] Relief Granted - Withholding of Deportation / Removal",
                    "[5D] Final Order of Deportation / Removal - Deferred Action Granted",
                    "[5E] Relief Granted - Extended Voluntary Departure",
                    "[5F] Unable to Obtain Travel Document",
                    "[8A] Excludable / Inadmissible - Hearing Not Commenced",
                    "[8B] Excludable / Inadmissible - Under Adjudication by IJ",
                    "[8C] Excludable / Inadmissible - Administrative Final Order Issued",
                    "[8D] Excludable / Inadmissible - Under Adjudication by BIA",
                    "[8E] Inadmissible - ICE Fugitive",
                    "[8F] Expedited Removal",
                    "[8G] Expedited Removal - Credible Fear Referral",
                    "[8H] Expedited Removal - Status Claim Referral",
                    "[8I] Inadmissible - ICE Fugitive - Expedited Removal",
                    "[8K] Expedited Removal Terminated due to Credible Fear Finding / NTA Issued",
                    "[9] VR Under Safeguards",
                    "[10] Visa Waiver Deportation / Removal",
                    "[11] Administrative Deportation / Removal",
                    "[12] Judicial Deportation / Removal",
                    "[13] Section 250 Removal",
                    "[14] Crewmen, Stowaways, S-Visa Holders, 235(c) Cases",
                    "[15] Terrorist Court Case (Title 5)",
                    "[16] Reinstated Final Order")

removals_df <- removals_df %>% 
  mutate(case_category = factor(case_category, ordered=TRUE, levels=category_order))

arrest_df <- arrest_df %>% 
  mutate(case_category = factor(case_category, ordered=TRUE, levels=category_order))

```

Here we examine the relationship of unique identifiers across datasets for arrest records with associated removals during 2025 (indicated by `departed_date` field) and removals records (by definition, removals occurring in 2025). A majority of unique identifier values are present in both datasets, indicating removals proceeding from ICE arrests. A sizeable minority are present in removals dataset only (likely indicating removals proceeding from arrests by CBP, not ICE; or from arrests which occurred prior to start of arrest dataset). A very small fraction are present in the arrests dataset only, likely indicating noise/errors in the data.

```{r plot_id_time}

arrests_removed_2025 <- latest_arrest %>% 
  filter(departed_date > "2024-12-31",
         !is.na(departed_date))

removal_ids <- unique(removals_df$unique_identifier)
arrest_ids <- unique(arrests_removed_2025$unique_identifier)

sets <- list("arrests" = arrest_ids,
             "removals" = removal_ids)

vd1 <- ggVennDiagram(x = sets,
              force_upset = TRUE, order.set.by = "name", order.intersect.by = "size"
              )

vd1 

```

There is no obvious common pattern for small subset of arrest records indicating 2025 deportation which do not appear in removals dataset.

```{r data_check}

arrests_removed_2025 <- arrests_removed_2025 %>% 
  mutate(id_in_removals = unique_identifier %in% unique(removals_df$unique_identifier))

data_check <- arrests_removed_2025 %>%
  filter(id_in_removals == FALSE)

```

Trends for total arrests and removals per month. Note shorter timeline of removals data compared to arrests data. Rates of removals consistently outpace rate of arrests.

```{r national_arrests_removals}

dat1 <- arrest_df %>% 
  mutate(month = floor_date(apprehension_date, "month")) %>% 
  count(month) %>% 
  mutate(type = "arrests")

dat2 <- removals_df %>% 
  mutate(month = floor_date(departed_date, "month")) %>% 
  count(month) %>% 
  mutate(type = "removals")

dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  filter(
         month < "2025-06-01") %>%
  ggplot(aes(x=month, y=n, color=type)) +
  geom_line() +
  ylim(0, NA) +
  labs(title = "National monthly ICE arrests and removals",
       caption = "Source: ICE data obtained by Deportation Data Project")

p1

```

# Analysis of `case_category` trends

`case_category` values suggest disposition of immigration case at time of generation of record; see [Kerwin et al. (2015)](https://doi.org/10.1177/233150241500300402) for interpretation of codes.

First we observe `case_category` trends for removals data. The data suggests that increase in removals from Jan. 2025 is primarily driven by people with final orders of removal, though expedited removals and reinstatements of removal trend up from April-May.

```{r case_category_trend_removals}

dat <- removals_df %>% 
  mutate(month = floor_date(departed_date, "month")) %>% 
  count(month, case_category)
  
p1 <- dat %>% 
  filter(month >= "2023-11-01",
         month < "2025-05-01") %>% 
  ggplot(aes(x = month, y = n, color = case_category)) +
  geom_line() +
  labs(title = "Removals by `case_category`")

plotly::ggplotly(p1)

```

We can also examine `case_category` trends for arrest records that include fields denoting a deportation. Note that this will only be representative of removals proceeding from ICE arrests. While specific trends differ (note different timeline for intersection of categories "[8C]" and "[16]" compared to above), overall trend is similar: increase in deportations of people with final orders; as well as increase in expedited removals and reinstatement of removal.

```{r case_category_trend_arrests}

top_case_category_arrest <- arrest_df %>% 
  count(case_category) %>% 
  arrange(desc(n))

dat <- latest_arrest %>% 
  filter(!is.na(departed_date)) %>% 
  mutate(month = floor_date(departed_date, "month"),
         # case_category = case_when(case_category %in% head(top_case_category_arrest$case_category) ~ case_category,
         #                           TRUE ~ "All others")
         ) %>% 
  count(month, case_category)
  
p1 <- dat %>% 
  filter(month >= "2023-10-01",
         month < "2025-05-01") %>% 
  ggplot(aes(x = month, y = n, color = case_category)) +
  geom_line() +
  labs(title = "Arrests with removal data")

plotly::ggplotly(p1)

```

## Seattle AOR trends

Here we observe `case_category` trends for Seattle Area of Responsibility arrests with associated removal data; these are similar to national trends above.

```{r}

dat <- arrest_df %>% 
  filter(!is.na(departed_date),
         apprehension_aor == "Seattle Area of Responsibility") %>% 
  mutate(month = floor_date(departed_date, "month")) %>% 
  count(month, case_category)
  
p1 <- dat %>% 
  filter(month >= "2023-10-01",
         month < "2025-05-01") %>% 
  ggplot(aes(x = month, y = n, color = case_category)) +
  geom_line()

plotly::ggplotly(p1)

```

Here we observe `case_category` trends for removals with `docket_aor` equal to "Seattle Area of Responsibility"; these are similar to national trends above. See below for analysis of relationships between `apprehension_aor` and `docket_aor`; in majority of cases these coincide.

```{r}

sea_top_categories <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  count(case_category) %>% 
  arrange(desc(n))

dat <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(month = floor_date(departed_date, "month"),
         case_category = case_when(case_category %in% head(sea_top_categories$case_category) ~ case_category,
                                   TRUE ~ "All others")) %>% 
  count(month, case_category)
  
p1 <- dat %>% 
  filter(month >= "2023-11-01",
         month < "2025-05-01") %>% 
  ggplot(aes(x = month, y = n, color = case_category)) +
  geom_line()

plotly::ggplotly(p1)

```

# Rights implications of `case_category` codes

`case_category` codes can be tentatively parsed for implications for human rights and due process. Some categories (listed below) may indicate that individuals have pending cases before U.S. immigration courts, and therefore should not be subjected to deportation. Other categories of removals, including administrative deportations, expedited removal, and reinstatement of removal may not involve judicial review; or include categories indicating asylum claims that should block removal.

```{r}

pending <- c('2A', '2B', '8A', '8B', '8D')
benefit <- c('5C', '5D')
expedited_admin_rein <- c('8F', '8G', '8H', '8I', '11', '16')

dat <- removals_df %>% 
  mutate(month = floor_date(departed_date, "month"),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", ""),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(month, docket_aor, case_category_group) %>% 
  group_by(month, docket_aor) %>% 
  mutate(pct = n / sum(n))
  
p1 <- dat %>% 
  filter(month >= "2023-11-01",
         month < "2025-06-01") %>% 
  ggplot(aes(x = month, y = n, fill = case_category_group)) +
  geom_col() +
  facet_wrap(~docket_aor, scales="free_y")

plotly::ggplotly(p1)

```

```{r dat_sea_basic}

dat_sea <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(case_category_group)

```

```{r}

dat_natl <- removals_df %>% 
  mutate(month = floor_date(departed_date, "month"),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", ""),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(month, case_category_group) %>% 
  group_by(month) %>% 
  mutate(pct = n / sum(n),
         type = "natl")

dat_sea <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(month = floor_date(departed_date, "month"),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(month, case_category_group) %>% 
  group_by(month) %>% 
  mutate(pct = n / sum(n),
         type = "sea")

dat <- rbind(dat_natl, dat_sea)
  
p1 <- dat %>% 
  filter(month >= "2023-11-01",
         month < "2025-06-01") %>% 
  ggplot(aes(x = month, y = n, fill = case_category_group)) +
  geom_col() +
  facet_wrap(~type, scales="free_y")

plotly::ggplotly(p1)

```

```{r}

dat_natl <- arrest_df %>% 
    filter(!is.na(departed_date),
          departed_date > "2023-10-31") %>% 
  mutate(month = floor_date(departed_date, "month")) %>% 
  count(month, case_category) %>% 
  group_by(month) %>% 
  mutate(pct = n / sum(n),
         type = "natl")

dat_sea <- arrest_df %>% 
    filter(!is.na(departed_date),
          departed_date > "2023-10-31") %>% 
  filter(apprehension_aor == "Seattle Area of Responsibility") %>% 
  mutate(month = floor_date(departed_date, "month")) %>% 
  count(month, case_category) %>% 
  group_by(month) %>% 
  mutate(pct = n / sum(n),
         type = "sea")

dat <- rbind(dat_natl, dat_sea)
  
p1 <- dat %>% 
  filter(month >= "2023-11-01",
         month < "2025-06-01") %>% 
  ggplot(aes(x = month, y = n, fill = case_category)) +
  geom_col() +
  facet_wrap(~type, scales = "free_y")

plotly::ggplotly(p1)


```

```{r}

dat_natl <- arrest_df %>% 
    filter(!is.na(departed_date),
          departed_date > "2023-10-31") %>% 
  mutate(month = floor_date(departed_date, "month"),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(month, case_category_group) %>% 
  group_by(month) %>% 
  mutate(pct = n / sum(n),
         type = "natl")

dat_sea <- arrest_df %>% 
    filter(!is.na(departed_date),
          departed_date > "2023-10-31") %>% 
  filter(apprehension_aor == "Seattle Area of Responsibility") %>% 
  mutate(month = floor_date(departed_date, "month"),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(month, case_category_group) %>% 
  group_by(month) %>% 
  mutate(pct = n / sum(n),
         type = "sea")

dat <- rbind(dat_natl, dat_sea)
  
p1 <- dat %>% 
  filter(month >= "2023-11-01",
         month < "2025-06-01") %>% 
  ggplot(aes(x = month, y = n, fill = case_category_group)) +
  geom_col() +
  facet_wrap(~type, scales = "free_y")

plotly::ggplotly(p1)


```

```{r}

dat_natl <- arrest_df %>% 
  filter(!is.na(departed_date),
          departed_date > "2023-10-31") %>% 
  mutate(admin = case_when(departed_date >= "2025-01-20" ~ "Trump",
                           departed_date < "2025-10-20" ~ "Biden"),
         month = floor_date(departed_date, "month"),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(admin, case_category, case_category_group) %>% 
  group_by(admin) %>%
  mutate(pct = n / sum(n),
         type = "natl")

dat_sea <- arrest_df %>% 
    filter(!is.na(departed_date),
          departed_date > "2023-10-31",
          apprehension_aor == "Seattle Area of Responsibility") %>% 
  mutate(admin = case_when(departed_date >= "2025-01-20" ~ "Trump",
                           departed_date < "2025-10-20" ~ "Biden"),
         month = floor_date(departed_date, "month"),
         case_category_group = case_when(str_detect(case_category, paste0(pending, collapse="|")) ~ "pending case",
                                         str_detect(case_category, paste0(benefit, collapse="|")) ~ "benefit blocking deport",
                                         str_detect(case_category, paste0(expedited_admin_rein, collapse="|")) ~ "lack of due process",
                                         TRUE ~ "all other categories"
                                         )) %>% 
  count(admin, case_category, case_category_group) %>% 
  group_by(admin) %>%
  mutate(pct = n / sum(n),
         type = "sea")

dat <- rbind(dat_natl, dat_sea)

plotly::ggplotly(p1)

# Biden admin period
d1 <- as.Date("2023-11-01")    
d2 <- as.Date("2025-01-19")
# calculation 
biden_months <- as.numeric((d2 - d1)/(365.25/12))

# Trump admin period
d1 <- as.Date("2025-01-20")    
d2 <- as.Date("2025-06-11")
# calculation 
trump_months <- as.numeric((d2 - d1)/(365.25/12))

tab <- dat %>% 
  pivot_wider(id_cols = c(case_category, case_category_group), names_from = c(admin, type), names_sep = "_", values_from = c(pct, n)) %>% 
  mutate(mean_Biden_natl = n_Biden_natl / biden_months,
         mean_Trump_natl = n_Trump_natl / trump_months,
         mean_Biden_sea = n_Biden_sea / biden_months,
         mean_Trump_sea = n_Trump_sea / trump_months)

tab2 <- tab %>% 
  group_by(case_category_group) %>% 
  summarize(pct_Biden_natl = sum (pct_Biden_natl, na.rm=TRUE),
            pct_Trump_natl = sum (pct_Trump_natl, na.rm=TRUE),
            n_Biden_natl = sum (n_Biden_natl, na.rm=TRUE),
            n_Trump_natl = sum (n_Trump_natl, na.rm=TRUE),
            mean_Biden_natl = sum (mean_Biden_natl, na.rm=TRUE),
            mean_Trump_natl = sum (mean_Trump_natl, na.rm=TRUE),
            pct_Biden_sea = sum (pct_Biden_sea, na.rm=TRUE),
            pct_Trump_sea = sum (pct_Trump_sea, na.rm=TRUE),
            n_Biden_sea = sum (n_Biden_sea, na.rm=TRUE),
            n_Trump_sea = sum (n_Trump_sea, na.rm=TRUE),
            mean_Biden_sea = sum (mean_Biden_sea, na.rm=TRUE),
            mean_Trump_sea = sum (mean_Trump_sea, na.rm=TRUE)
            )

write_csv(tab, here('analyze', 'output', 'case_category_tab.csv'), na = "")

write_csv(tab2, here('analyze', 'output', 'case_category_tab2.csv'), na = "")
```


# Arrests X removals comparison

Minority (but not insignificant quantity) of unique ids appearing in both arrests and removals have inconsistent `case_status` and/or `case_category` values. This might not be unexpected if arrests and removals datasets were generated at different times.

```{r}

arrests_subset <- arrest_df %>% 
  dplyr::select(unique_identifier, case_status, case_category, n_dupe)

removals_subset <- removals_df %>% 
  dplyr::select(unique_identifier, case_status, case_category, n_dupe)

compare_subsets <- inner_join(arrests_subset, removals_subset, by="unique_identifier", suffix = c(".arrests", ".removals"), relationship = "many-to-many")

compare_subsets$status_equal <- compare_subsets$case_status.arrests == compare_subsets$case_status.removals

compare_subsets$category_equal <- compare_subsets$case_category.arrests == compare_subsets$case_category.removals

unequal_compare <- compare_subsets %>% 
  filter(status_equal != TRUE | category_equal != TRUE)

unequal_status <- compare_subsets %>% 
  filter(status_equal != TRUE)

unequal_category <- compare_subsets %>% 
  filter(category_equal != TRUE)

```

No obvious relationship between arrest `apprehension_method` and removal `final_program`; this suggests `final_program` is not analogous to "arresting agency", at least in removals dataset.

```{r}

arrests_subset <- latest_arrest %>% 
  dplyr::select(unique_identifier, apprehension_method)

removals_subset <- latest_removal %>% 
  dplyr::select(unique_identifier, final_program)

dat <- inner_join(arrests_subset, removals_subset, by="unique_identifier", suffix = c(".arrests", ".removals"))

t1 <- dat %>% 
  count(apprehension_method, final_program)


```

Usually consistent relationship between `apprehension_aor`, `docket_aor`.

```{r}

arrests_subset <- latest_arrest %>% 
  dplyr::select(unique_identifier, apprehension_aor)

removals_subset <- latest_removal %>% 
  dplyr::select(unique_identifier, docket_aor)

dat <- inner_join(arrests_subset, removals_subset, by="unique_identifier", suffix = c(".arrests", ".removals"))

t1 <- dat %>% 
  count(apprehension_aor, docket_aor)


```

Inconsistent relationship between `final_program` values in arrests/removals

```{r}

arrests_subset <- latest_arrest %>% 
  dplyr::select(unique_identifier, final_program)

removals_subset <- latest_removal %>% 
  dplyr::select(unique_identifier, final_program)

dat <- inner_join(arrests_subset, removals_subset, by="unique_identifier", suffix = c(".arrests", ".removals"))

t1 <- dat %>% 
  count(final_program.arrests, final_program.removals)


```

```{r}

arrests_subset <- latest_arrest %>% 
  dplyr::select(unique_identifier, departed_date) %>% 
  mutate(departed_date = as.Date(departed_date))

removals_subset <- latest_removal %>% 
  dplyr::select(unique_identifier, departed_date) %>% 
  mutate(departed_date = as.Date(departed_date))


dat <- inner_join(arrests_subset, removals_subset, by="unique_identifier", suffix = c(".arrests", ".removals")) %>% 
  filter(!is.na(departed_date.arrests)) %>% 
  mutate(unequal_departed_date = departed_date.arrests != departed_date.removals) 




```