---
title: "Comparison of anonymized identifiers across datasets"
author: "UWCHR"
date: "2025-03-31"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---


```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram, UpSetR)

```


```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Encounters_LESA-STU_FINAL_Redacted_raw.xlsx"

encounter_df_1 <- read_excel(here('analyze', 'input', file), sheet=1, skip=6) %>%
  janitor::clean_names() %>%
  mutate(event_date = as.Date(event_date))

encounter_df_2 <- read_excel(here('analyze', 'input', file), sheet=2, skip=6) %>%
  janitor::clean_names() %>%
  mutate(event_date = as.Date(event_date))

encounter_df <- rbind(encounter_df_1, encounter_df_2)

rm(encounter_df_1, encounter_df_2)

encounter_ids <- unique(encounter_df$unique_identifier)

```


```{r setup_detainers}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Detainers_LESA-STU_FINAL_Redacted_raw.xlsx"

detainer_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names()

detainer_df<- detainer_df%>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units = "days"))

detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Admin_Arrests_LESA-STU_FINAL_Redacted_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names()

arrest_df <- arrest_df %>% 
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("unique_identifier")) |>
  filter(!is.na(unique_identifier))

arrest_ids <- unique(arrest_df$unique_identifier)

```

```{r setup_detention, message=FALSE, warning=FALSE, include=TRUE}

file <- "ice_detentions_nov23-jul25.csv.gz"

detention_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         stay_book_out_date = ymd(stay_book_out_date),
         book_in_fy = substr(quarter(book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
         stay_book_in_fy = substr(quarter(stay_book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4)) %>% filter(!is.na(unique_identifier))


detention_df <- detention_df %>% 
  group_by(unique_identifier) %>% 
  arrange(stay_book_in_date_time, book_in_date_time, unique_identifier) %>% 
  ungroup()

detention_ids <- unique(detention_df$unique_identifier)

```

```{r setup_removals, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ICE_Removals07312025.xlsx"

removal_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% filter(!is.na(unique_identifier))

removal_ids <- unique(removal_df$unique_identifier)

```

```{r compare_sets}

data_levels <- c("encounter", "detainer", "arrest", "detention", "removal")

sets <- list("encounters" = encounter_ids,
             "detainers" = detainer_ids,
             "arrests" = arrest_ids,
             "detentions" = detention_ids,
             "removals" = removal_ids)

# p1 <- ggVennDiagram(x = sets,
#                     color = name,
#                     label = "count") +
#   scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")
# 
# p1

p2 <- ggVennDiagram(x = sets,
              force_upset = TRUE, order.set.by = "name", order.intersect.by = "size"
              )

p2 

```

```{r plot_all_enforcement}

dat1 <- encounter_df %>% 
  filter(event_date >= "2023-11-01",
    event_date <= "2025-05-30") %>% 
  mutate(yearmon = as.yearmon(event_date)) %>% 
  count(yearmon) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(detainer_prepare_date >= "2023-11-01",
         detainer_prepare_date <= "2025-05-30") %>% 
  mutate(yearmon = as.yearmon(detainer_prepare_date)) %>% 
  count(yearmon) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(apprehension_date >= "2023-11-01",
    apprehension_date <= "2025-05-30") %>% 
  mutate(yearmon = as.yearmon(apprehension_date)) %>% 
  count(yearmon) %>% 
  mutate(type = "arrest")

dat4 <- removal_df %>% 
  filter(departed_date >= "2023-11-01",
         departed_date <= "2025-05-30") %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  count(yearmon) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4)

p1 <- dat %>% 
  ggplot(aes(x = yearmon, y = n, fill = type)) +
  geom_col(aes(group = type), position="dodge") +
  ylim(0, NA)

p1

```

https://ohss.dhs.gov/khsm/ice-detentions

```{r plot_detention_book_in_out}

dat5 <- detention_df %>% 
  filter(stay_book_in_date_time >= "2023-11-01",
       book_in_date_time >= "2023-11-01",
       book_in_date_time <= "2025-05-30",
       placement_count == 1,
       ) %>% 
    mutate(yearmon = as.yearmon(book_in_date_time),
           book_in_year = year(stay_book_in_date_time)) %>% 
  count(yearmon) %>% 
  mutate(type = "initial detention book-in")

dat6 <- detention_df %>% 
  filter(stay_book_out_date_time >= "2023-11-01",
       detention_book_out_date_time  >= "2023-11-01",
       detention_book_out_date_time <= "2025-05-30",
       placement_count == total_placements,
       ) %>% 
    mutate(yearmon = as.yearmon(detention_book_out_date_time),
           book_in_year = year(detention_book_out_date_time)) %>% 
  count(yearmon) %>% 
  mutate(type = "detention book-out")

dat <- rbind(dat5, dat6)

p1 <- dat %>% 
  ggplot(aes(x = yearmon, y = n, fill = type)) +
  geom_col(aes(group = type), position="dodge") +
  ylim(0, NA)

p1

```

Appearance of `unique_identifier` values within and across datasets. What can we do with this? No reason not to expect that any of these datasets could have repeat entries per individual.

```{r compare_n_id}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  count(unique_identifier) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  count(unique_identifier) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  count(unique_identifier) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  count(unique_identifier) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  count(unique_identifier) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier'), names_from = "type", values_from = "n")

```

A small subset of meaningful fields are shared by all datasets. Other fields shared by two or more datasets, could analyze these as well.

```{r shared_cols}

encounter_cols <- names(encounter_df)
detainer_cols <- names(detainer_df)
arrest_cols <- names(arrest_df)
detention_cols <- names(detention_df)
removal_cols <- names(removal_df)

shared_cols <- Reduce(intersect, list(encounter_cols, detainer_cols, arrest_cols, detention_cols, removal_cols))

```

This is a crude comparison of values per unique identifier across datasets. Highlights cases where specific values are perturbed across datasets. Would we expect these values to be constant across datasets, presuming generated at same time?

Crude because `distinct()` naively selects first occurrence of `unique_identifier` per dataset, when repeat entries per `unique_identifier` could be associated with perturbation of values across datasets. Better to check for perturbation across `unique_identifier` within datasets first? Or restrict to `unique_identifier` values with or without repeat records (not realistic for detentions data).

```{r compare_gender}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, gender) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, gender) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, gender) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, gender) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, gender) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat <- dat %>% 
  group_by(unique_identifier) %>% 
  mutate(n_dist = n_distinct(gender, na.rm=TRUE)) %>% 
  ungroup()

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier', 'n_dist'), names_from = "type", values_from = "gender") %>% 
  filter(n_dist > 1)

p1 <- dat %>% 
  count(gender, type) %>% 
  mutate(type = factor(type, levels=data_levels)) %>% 
  ggplot(aes(x = type, y = n, fill = gender)) +
  geom_col()

p1

```

```{r compare_citizenship}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat <- dat %>% 
  group_by(unique_identifier) %>% 
  mutate(n_dist = n_distinct(citizenship_country, na.rm=TRUE))

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier', 'n_dist'), names_from = "type", values_from = "citizenship_country") %>% 
  filter(n_dist > 1)



```

```{r compare_departed}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, citizenship_country) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat <- dat %>% 
  group_by(unique_identifier) %>% 
  mutate(n_dist = n_distinct(citizenship_country, na.rm=TRUE))

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier', 'n_dist'), names_from = "type", values_from = "citizenship_country") %>% 
  filter(n_dist > 1)

```

```{r compare_final_order_date}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_order_date) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_order_date) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_order_date) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_order_date) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_order_date) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat <- dat %>% 
  group_by(unique_identifier) %>% 
  mutate(n_dist = n_distinct(final_order_date, na.rm=TRUE))

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier', 'n_dist'), names_from = "type", values_from = "final_order_date") %>% 
  filter(n_dist > 1)

```

```{r compare_final_program}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_program) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_program) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_program) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_program) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, final_program) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat <- dat %>% 
  group_by(unique_identifier) %>% 
  mutate(n_dist = n_distinct(final_program, na.rm=TRUE)) %>% 
  ungroup()

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier', 'n_dist'), names_from = "type", values_from = "final_program") %>% 
  filter(n_dist > 1)

p1 <- dat %>% 
  count(final_program, type) %>% 
  mutate(type = factor(type, levels=data_levels)) %>% 
  ggplot(aes(x = type, y = n, fill = final_program)) +
  geom_col()

plotly::ggplotly(p1)

p1.1 <- dat %>% 
  count(final_program, type) %>% 
  mutate(type = factor(type, levels=data_levels)) %>% 
  ggplot(aes(x = type, y = n, fill = final_program)) +
  geom_col(position = "fill")

plotly::ggplotly(p1.1)

```

```{r compare_case_category}

dat1 <- encounter_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, case_category) %>% 
  mutate(type = "encounter")

dat2 <- detainer_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, case_category) %>% 
  mutate(type = "detainer")

dat3 <- arrest_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, case_category) %>% 
  mutate(type = "arrest")

dat4 <- detention_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, case_category) %>% 
  mutate(type = "detention")

dat5 <- removal_df %>% 
  filter(!is.na(unique_identifier)) %>% 
  distinct(unique_identifier, .keep_all = TRUE) %>% 
  dplyr::select(unique_identifier, case_category) %>% 
  mutate(type = "removal")

dat <- rbind(dat1, dat2, dat3, dat4, dat5)

dat <- dat %>% 
  group_by(unique_identifier) %>% 
  mutate(n_dist = n_distinct(case_category, na.rm=TRUE)) %>% 
  ungroup()

p1 <- dat %>%
  count(type, case_category) %>% 
  mutate(type = factor(type, levels = data_levels)) %>% 
  ggplot(aes(x = type, y = n, fill = case_category)) +
  geom_col(position = "fill")

plotly::ggplotly(p1)

dat_wide <- pivot_wider(dat, id_cols = c('unique_identifier', 'n_dist'), names_from = "type", values_from = "case_category") %>% 
  filter(n_dist > 1)


```