---
title: "Analysis of Detainers of Interest"
author: "UWCHR"
date: "2025-06-07"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram)

```

```{r setup_facilities, message=FALSE, warning=FALSE, include=TRUE}

# Running into problem with .gz version of facilities dataset
file <- "facilities.csv"

facilities_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names()

```

```{r setup_detainers, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Detainers_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detainer_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units='days'),
         pre_trump = detainer_prepare_date < as.Date("2025-01-20"))

detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Encounters_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

encounter_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date),
         pre_trump = event_date < as.Date("2025-01-20"))

encounter_ids <- unique(encounter_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Admin_Arrests_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(apprehension_date = as.Date(apprehension_date),
         apprehension_method = fct(apprehension_method),
         criminality = apprehension_criminality == "1 Convicted Criminal",
         n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier"),
         pre_trump = apprehension_date < as.Date("2025-01-20")) %>% 
  filter(is.na(unique_identifier) | n_dupe == 1)

arrest_ids <- unique(arrest_df$unique_identifier)

```

```{r setup_detentions, message=FALSE, warning=FALSE, include=TRUE}

file <- "ice_detentions_nov23-feb25.csv.gz"

detain_unique_stays_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names() %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         pre_trump = stay_book_in_date_time < as.Date("2025-01-20")) %>% 
  left_join(facilities_df, by=c("detention_facility_code" = "detloc"))

```

```{r}
# Break down datasets by geography and pre vs. post Trump
segment_df <- function(df, geo_filter) {
  df_wa <- geo_filter(df)
  df_pre_trump <- filter(df, pre_trump == TRUE)
  df_post_trump <- filter(df, pre_trump == FALSE)
  df_pre_trump_wa <- geo_filter(filter(df, pre_trump == TRUE))
  df_post_trump_wa <- geo_filter(filter(df, pre_trump == FALSE))
  
  return (list("wa"=df_wa,
            "pre_trump"=df_pre_trump,
            "post_trump"=df_post_trump,
            "pre_trump_wa"=df_pre_trump_wa,
            "post_trump_wa"=df_post_trump_wa))
}

detainer_dfs <- segment_df(detainer_df,
                           (function(df) return(filter(df, facility_state == "WASHINGTON"))))

detention_dfs <- segment_df(detention_df,
                            (function(df) return(filter(df, state == "WA"))))

encounter_dfs <- segment_df(encounter_df,
                            (function(df) return(filter(df, responsible_aor == "Seattle Area of Responsibility"))))

arrest_dfs <- segment_df(arrest_df,
                         (function(df) return(filter(df, apprehension_aor == "Seattle Area of Responsibility"))))
```

```{r}
# Convert decimal to percentage
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}
```

```{r}
time_filter <- 2  # date_diff.detainer_arrest threshold (in days)

detainer_arrest_wa_df <- inner_join(detainer_dfs[["wa"]], arrest_df,
                                    by="unique_identifier",
                                    suffix=c(".detainer", ".arrest"),
                                    na_matches='never',
                                    relationship='many-to-many') %>% 
  mutate(date_diff.detainer_arrest = difftime(apprehension_date.arrest, detainer_prepare_date, units="days")) %>%
  filter(date_diff.detainer_arrest >= 0) %>%
  filter(!(detention_facility %in% c("SEATAC FED.DET.CENTER", "STATE PRISON, SHELTON", "US MARSHALS, YAKIMA", "US MARSHALS, SPOKANE,WA", "WASHINGTON STATE CORRECTIONS", "STAFFORD CREEK CORRECTIONS"))) %>%
  mutate(possible_kww_violation = date_diff.detainer_arrest <= time_filter)
```

The following charts show the frequency and dates of detainers sharing unique identifiers with arrests, without filtering by time.
```{r}
# No time filter
ggplot(detainer_arrest_df_wa, aes(x=detention_facility.detainer)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1) +
  coord_cartesian(ylim = c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Number of Detainers Sharing Unique Identifiers with Arrests, by Facility (No Time Filter)")

ggplot(detainer_arrest_df_wa, aes(x=detainer_prepare_date.detainer)) +
  geom_dotplot() +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Date Prepared of Detainers Sharing Unique Identifiers with Arrests (No Time Filter)")
```

The following charts show the frequency and dates of ``detainers of interest,'' or detainers sharing unique identifiers with arrests and placed within `r time_filter` days of an arrest. 
```{r}
# With time filter
detainers_of_interest <- detainer_arrest_df_wa %>% 
  filter(possible_kww_violation == TRUE)

ggplot(detainers_of_interest, aes(x=detention_facility.detainer)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1) +
  coord_cartesian(ylim = c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Number of Detainers Sharing Unique Identifiers with Arrests by Facility (Time Filter)")

ggplot(detainers_of_interest, aes(x=detainer_prepare_date.detainer)) +
  geom_dotplot() +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Date Prepared of Detainers Sharing Unique Identifiers with Arrests (Time Filter)")
```

In total, there are `r nrow(detainers_of_interest)` detainers of interest, as shown in the following table.
```{r}
detainers_of_interest[c("unique_identifier", "detainer_prepare_date.detainer", "apprehension_date.arrest", "date_diff.detainer_arrest", "detention_facility.detainer")]
print(nrow(detainers_of_interest))
```

The following table displays the average time between detainers and arrests with shared unique identifiers by detention facility.
```{r}
# Average time per facility
date_diff_by_facility <- detainer_arrest_df_wa %>% 
  group_by(detention_facility.detainer) %>% 
  summarize(mean=mean(date_diff.detainer_arrest))
date_diff_by_facility
```