---
title: "ICE ERO-LESA detention data Oct. 2022 - Mar. 2026"
author: "UWCHR"
date: "2026-04-01"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---


```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, digest, ggsankeyfier)

arrests <- arrow::read_parquet("https://github.com/deportationdata/ice/raw/refs/heads/main/data/arrests-latest.parquet")

detentions <- arrow::read_parquet("https://github.com/deportationdata/ice/raw/refs/heads/main/data/detention-stints-latest.parquet")

headcount <- arrow::read_parquet("https://github.com/deportationdata/ice/raw/refs/heads/main/data/facilities-daily-population-latest.parquet")

pnw_facil <- c("FDLHOLD",
"HAVMCWA",
"RICHOLD",
"SPOHOLD",
"STJMCWA",
"TACGHWA",
"WNTHOLD",
"ABRDNWA",
"ADAMSWA",
"BLHHOLD",
"BENTOWA",
"CHELAWA",
"CLJUVWA",
"CLARKWA",
"COWLIWA",
"COWJVWA",
"ECOLOWA",
"FERRYWA",
"FORKSWA",
"FRANKWA",
"GRANTWA",
"KINCOWA",
"KITTIWA",
"MARTIWA",
"MASONWA",
"WAMCNEI",
"CSCNWWA",
"OKANOWA",
"PENDOWA",
"RRFINWA",
"REGJCWA",
"BOPSET",
"SEAICDF",
"SEAHOLD",
"SPOKAWA",
"STEVEWA",
"SUNNYWA",
"THURSWA",
"WHICOWA",
"YIKIMWA",
"YAKHOLD",
"MEDHOLD",
"POOHOLD",
"EUGHOLD")

```

```{r detain_setup, message=FALSE, warning=FALSE, include=TRUE}

max_date <- max(detentions$book_in_date_time, na.rm=TRUE)

detentions <- detentions %>% 
  filter(!is.na(stay_book_in_date_time),
         !is.na(book_in_date_time)) %>% 
  mutate(stay_length = difftime(stay_book_out_date_time,
            stay_book_in_date_time, unit='days'),
         placement_length = difftime(book_out_date_time,
            book_in_date_time, unit='days'),
         stay_length_elapsed = difftime(replace_na(stay_book_out_date_time, max_date),
            stay_book_in_date_time, unit='days'),
         placement_length_elapsed = difftime(replace_na(book_out_date_time, max_date),
            book_in_date_time, unit='days'),
         initial_placement_exact = stay_book_in_date_time == book_in_date_time,
         initial_placement_approx = date(stay_book_in_date_time) == date(book_in_date_time)
         )

detentions <- detentions %>% 
  filter(!is.na(unique_identifier)) %>% 
  group_by(unique_identifier) %>% 
  arrange(stay_book_in_date_time, book_in_date_time) %>%
  mutate(total_stays = n_distinct(stay_book_in_date_time),
         total_placements = n(),
         current_stay = is.na(stay_book_out_date_time),
         current_placement = is.na(book_out_date_time),
         stay_count = data.table::rleid(stay_book_in_date_time)) %>% 
  ungroup()

detentions <- detentions %>% 
  group_by(stay_ID) %>% 
  arrange(stay_book_in_date_time, book_in_date_time) %>%
  mutate(placement_count = data.table::rleid(stint_ID),
         stay_placements = n(),
         first_facil = detention_facility_code[[1]], 
         last_facil = detention_facility_code[[length(detention_facility_code)]],
         prev_facil = lag(detention_facility_code, n=1),
         next_facil = lead(detention_facility_code, n=1),
         longest_placement_facil = detention_facility_code[which.max(placement_length_elapsed)],
         last_placement = placement_count == stay_placements,
         longest_placement = placement_length_elapsed == max(placement_length_elapsed)
         ) %>% 
  ungroup()

```

```{r, message=FALSE, warning=FALSE, include=TRUE}

url <- "https://drive.usercontent.google.com/download?id=1XAOS7t3ijImVXQ9G7l6xdCr_rBh8i4cz&export=download&authuser=0&confirm=t&uuid=0c4a4b74-5e5f-4f00-8c8a-eb00a56708f9&at=APcXIO3JvUiU3ilszJZ3EAuIE01Y:1772065031530"
f <- tempfile(fileext = ".csv.gz")
download.file(url, f, mode = "wb")

uwchr <- read.delim(f, sep='|')

uwchr <- uwchr %>% 
  rename(apprehension_date = arrest_date) %>% 
  mutate(apprehension_date = mdy(apprehension_date),
    fy = as.numeric(substr(quarter(apprehension_date, fiscal_start = 10, with_year = TRUE), 0, 4)),
    fy_quarter = quarter(apprehension_date, fiscal_start = 10, with_year = TRUE),
    cy_quarter_zoo = as.yearqtr(apprehension_date, format = "%Y-%M-%D"))

arrests <- arrests %>% 
  mutate(fy = as.numeric(substr(quarter(apprehension_date, fiscal_start = 10, with_year = TRUE), 0, 4)),
         fy_quarter = quarter(apprehension_date, fiscal_start = 10, with_year = TRUE),
         cy_quarter_zoo = as.yearqtr(apprehension_date, format = "%Y-%M-%D"))

```

# Analysis

Average daily ICE arrests hit an all-time high during late 2025.

```{r}

dat1 <- arrests %>% 
  count(apprehension_date) %>% 
  mutate(moving_avg = rollmean(n, k = 90, fill = NA, align = "right"),
         source = "ddp")

dat2 <- uwchr %>% 
  count(apprehension_date) %>% 
  mutate(moving_avg = rollmean(n, k = 90, fill = NA, align = "right"), 
         source = "uwchr")

dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  ggplot(aes(x = apprehension_date, y = moving_avg, color = source)) +
  geom_line() +
  ylim(0, NA) +
  labs(title = "National daily ICE arrests, 90 day moving avg.")

p1

```

During the same time period, the nationwide detained population peaked at over 70,000; it declined slightly in early 2026.

```{r national_headcount}

uwchr_headcount_detloc <- read_delim(here::here('analyze', 'input', 'headcount_fy12-24ytd.csv.gz'), delim="|")

uwchr_headcount <- uwchr_headcount_detloc %>% 
  group_by(date) %>% 
  summarize(n_detained_at_midnight = sum(n)) %>% 
  mutate(moving_avg = rollmean(n_detained_at_midnight, k = 7, fill = NA, align = "right"),
         source= "uwchr")

ddp_headcount <- headcount %>% 
  group_by(date) %>% 
  summarize(n_detained_at_midnight = sum(n_detained_at_midnight)) %>% 
  mutate(moving_avg = rollmean(n_detained_at_midnight, k = 7, fill = NA, align = "right"),
         source = "ddp")


dat <- rbind(uwchr_headcount, ddp_headcount)

p1 <- dat %>% 
  filter(date >= "2011-10-01") %>% 
  ggplot(aes(x = date, y = moving_avg, color = source)) +
  geom_line() +
  ylim(0, NA) +
  labs(title= "National detained population, 7 day moving avg.")

p1

```

Seattle AOR arrests also hit an historic high point in late 2025.

```{r}

dat1 <- arrests %>% 
  filter(apprehension_aor == "Seattle Area of Responsibility") %>% 
  count(apprehension_date) %>% 
  mutate(moving_avg = rollmean(n, k = 90, fill = NA, align = "right"),
         source = "ddp")

dat2 <- uwchr %>% 
  filter(aor == "SEA") %>% 
  count(apprehension_date) %>% 
  mutate(moving_avg = rollmean(n, k = 90, fill = NA, align = "right"), 
         source = "uwchr")

dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  ggplot(aes(x = apprehension_date, y = moving_avg, color = source)) +
  geom_line() +
  ylim(0, NA) +
  labs(title = "Seattle AOR daily ICE arrests, 90 day moving avg.")

p1

```

Detained population at NWIPC dropped despite high rate of arrests. Why?

```{r cscnwwa_headcount}

cacnwwa_headcount_uwchr <- uwchr_headcount_detloc %>% 
  filter(detention_facility_code == "CSCNWWA") %>% 
  rename(n_detained_at_midnight = n) %>%
  mutate(moving_avg = rollmean(n_detained_at_midnight, k = 7, fill = NA, align = "right"),
         source = "uwchr") %>% 
  dplyr::select(date, n_detained_at_midnight, moving_avg, source)

cscnwwa_headcount_ddp <- headcount %>% 
  filter(detention_facility_code == "CSCNWWA") %>% 
  mutate(moving_avg = rollmean(n_detained_at_midnight, k = 7, fill = NA, align = "right"),
         source = "ddp") %>% 
  dplyr::select(date, n_detained_at_midnight, moving_avg, source)

dat <- rbind(cacnwwa_headcount_uwchr, cscnwwa_headcount_ddp)

p1 <- dat %>% 
  filter(date >= "2011-10-01") %>% 
  ggplot(aes(x = date, y = moving_avg, color = source)) +
  geom_line() +
  ylim(0, NA) +
  labs(title = "NWIPC daily population, 7 day moving avg.")

p1

```

Explanation 1: Slightly more people who are arrested by ICE are not detained

```{r}

arrests <- arrests %>% 
  filter(apprehension_date <= "2026-03-31") %>% 
  mutate(arrest_with_detention_history = unique_identifier %in% unique(detentions$unique_identifier),
         month = floor_date(apprehension_date, unit="month"))

dat <- arrests %>% 
  count(month, arrest_with_detention_history)

p1 <- dat %>% 
  ggplot(aes(x = month, y = n, color = arrest_with_detention_history)) +
  geom_line() + 
  ylim(0, NA) +
  labs(title= "National arrests per month",
       subtitle = "By whether associated with any detention history")

p1

dat <- arrests %>% 
  mutate(apprehension_aor = str_replace(apprehension_aor, " Area of Responsibility", "")) %>% 
  count(month, arrest_with_detention_history, apprehension_aor)

# p2 <- dat %>% 
#   ggplot(aes(x = month, y = n, color = arrest_with_detention_history)) +
#   geom_line() +
#   facet_wrap(~apprehension_aor) +
#   labs(title= "Arrests per month",
#        subtitle = "By whether associated with any detention history")
# 
# p2


p2 <- dat %>% 
  filter(apprehension_aor %in% c("Seattle")) %>% 
  ggplot(aes(x = month, y = n, color = arrest_with_detention_history)) +
  geom_line() +
  facet_wrap(~apprehension_aor) +
  labs(title= "Arrests per month",
     subtitle = "By whether associated with any detention history")

p2
```

Explanation 2: Increased detention "throughput"; more people are being booked in, but more also being booked out during same time. (Question: does this signal more "churn" overall or is there a portion of population that cycles thru versus portion of population which is warehoused?)

```{r}

cscnwwa_detentions <- detentions %>% 
  filter(detention_facility_code == "CSCNWWA")

cscnwwa_book_ins_weekly <- cscnwwa_detentions %>% 
  filter(book_in_date_time >= "2022-10-01") %>% 
  mutate(date = as.Date(book_in_date_time)) %>% 
  filter(!is.na(date)) %>% 
  mutate(week = floor_date(date, unit = "week", week_start = 1)) %>% 
  count(week) %>% 
  mutate(type = "bookins")

cscnwwa_book_outs_weekly <- cscnwwa_detentions %>% 
  filter(book_out_date_time >= "2022-10-01") %>% 
  mutate(date = as.Date(book_out_date_time)) %>% 
  filter(!is.na(date)) %>% 
  mutate(week = floor_date(date, unit = "week", week_start = 1)) %>% 
  count(week) %>% 
  mutate(type = "bookouts")

dat <- rbind(cscnwwa_book_ins_weekly, cscnwwa_book_outs_weekly)

p1 <- dat %>% 
  filter(week >= "2025-01-01") %>% 
  ggplot(aes(x = week, y = n, fill = type)) + 
  geom_col(position = "dodge") +
  labs(title = "NWIPC daily book-ins/book-outs")

p1

p2 <- dat %>% 
  filter(week >= "2025-01-01") %>% 
  ggplot(aes(x = week, y = n, fill = type)) + 
  geom_col(position = "stack") +
  labs(title = "NWIPC daily book-ins/book-outs (total throughput)")

p2

```

Most people booked out of NWIPC are transferred to other facilities; small but increasing proportion bonded out or released in various categories.

```{r}

cscnwwa_detentions <- cscnwwa_detentions %>% 
  mutate(book_in_week = floor_date(book_in_date_time, unit = "week", week_start = "Monday"),
    book_out_week = floor_date(book_out_date_time, unit = "week", week_start = "Monday"),
    book_in_month = floor_date(book_in_date_time, unit = "month"),
    book_out_month = floor_date(book_out_date_time, unit = "month"))

dat <- cscnwwa_detentions %>% 
  filter(book_out_date_time >= "2024-12-31") %>% 
  count(book_out_month, detention_release_reason)

p1 <- dat %>% 
  ggplot(aes(x = book_out_month, y = n, fill = detention_release_reason)) +
  geom_col()

ggplotly(p1)

```

Facilities people are transferred to has shifted over the year.

```{r}

dat <- cscnwwa_detentions %>% 
  filter(book_out_date_time >= "2024-12-31") %>% 
  filter(detention_release_reason == "Transferred") %>% 
  count(book_out_month, next_facil)

p1 <- dat %>% 
  ggplot(aes(x = book_out_month, y = n, fill = next_facil)) +
  geom_col()

ggplotly(p1)

```

People transferred to NWIPC from other facilities are primarily coming via local hold rooms.

```{r}

dat <- cscnwwa_detentions %>% 
  filter(book_in_date_time >= "2024-12-31") %>% 
  count(book_in_month, prev_facil)

p1 <- dat %>% 
  ggplot(aes(x = book_in_month, y = n, fill = prev_facil)) +
  geom_col() +
  gghighlight(prev_facil %in% pnw_facil)

p1

```
This closely tracks initial facility bookins for Seattle AOR.

```{r}

sea_bookins <- detentions %>% 
  filter(book_in_aor == "Seattle Area of Responsibility",
         book_in_date_time >= "2025-01-01") %>% 
  distinct(stay_ID, .keep_all = TRUE)

dat <- sea_bookins %>% 
  mutate(month = floor_date(book_in_date_time, "month")) %>% 
  count(month, detention_facility)

p1 <- dat %>% 
  ggplot(aes(x = month, y = n, fill = detention_facility)) +
  geom_col()

ggplotly(p1)

```

Transfers to Ft. Bliss/Camp East Montana make up a significant proportion of increased throughput during late 2025

```{r transfers_from_cscnwwa}

cscnwwa_transfers <- detentions %>% 
  filter(detention_facility_code == "CSCNWWA",
         detention_release_reason == "Transferred")

cscnwwa_transfer_destinations <- cscnwwa_transfers %>% 
  count(next_facil) %>% 
  arrange(desc(n))

p1 <- cscnwwa_transfers %>% 
  filter(book_out_date_time >= "2025-01-01") %>% 
  mutate(week = floor_date(book_out_date_time, "week", "Monday")) %>% 
  count(week, next_facil) %>% 
  ggplot(aes(x = week, y = n, fill = next_facil)) +
  geom_col() +
  gghighlight(next_facil == "EROFCB") +
  labs(title = "Weekly transfers from NWIPC",
         subtitle = "Transfers to ERO EL PASO CAMP EAST MONTANA highlighted")

p1

```

Majority transferred to EROFCB removed; 73 remained in detention at time of release of dataset.

```{r}

cscnwwa_to_erofcb <- detentions %>% 
  filter(detention_facility_code == "CSCNWWA",
         next_facil == "EROFCB")

p1 <- cscnwwa_to_erofcb %>%
  distinct(stay_ID, .keep_all = TRUE) %>% 
  count(stay_release_reason) %>% 
  ggplot(aes(x = "stay_release_reason", y = n, fill=stay_release_reason)) +
  geom_col(position="fill") +
  coord_polar(theta="y", start=0)

p1

```

```{r}

cscnwwa_transfers_2025 <- detentions %>% 
  distinct(stay_ID, .keep_all = TRUE) %>% 
  filter(book_out_date_time >= "2025-01-01",
         detention_facility_code == "CSCNWWA")

cscnwwa_destinations <- cscnwwa_transfers_2025 %>% 
  count(next_facil) %>% 
  arrange(desc(n))

dat <- cscnwwa_transfers_2025 %>% 
  mutate(next_facil = case_when(next_facil %in% head(cscnwwa_destinations$next_facil, 10) ~ next_facil,
         TRUE ~ "ALL OTHERS")) %>% 
  group_by(next_facil, stay_release_reason) %>% 
  summarize(n = n())

```

```{r}

cscnwwa_annual_book_ins <- detentions %>% 
  filter(book_in_date_time >= "2022-01-01", 
    detention_facility_code == "CSCNWWA") %>% 
  mutate(cy = year(book_in_date_time)) %>% 
  count(cy) %>% 
  mutate(type = "bookouts")

cscnwwa_annual_book_outs <- detentions %>% 
  filter(book_out_date_time >= "2022-01-01", 
    detention_facility_code == "CSCNWWA") %>% 
  mutate(cy = year(book_out_date_time)) %>% 
  count(cy) %>% 
  mutate(type = "bookins")

dat <- rbind(cscnwwa_annual_book_ins, cscnwwa_annual_book_outs)

p1 <- dat %>% 
  ggplot(aes(x = cy, y = n, fill = type)) +
  geom_col(position="dodge")

p1

```

