---
title: "Detention Flows Tabular Analysis"
author: "UWCHR"
date: "2025-06-07"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram)

```

```{r setup_facilities, message=FALSE, warning=FALSE, include=TRUE}

# Running into problem with .gz version of facilities dataset
file <- "facilities.csv"

facilities_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names()

```

```{r setup_detainers, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Detainers_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detainer_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units='days'),
         pre_trump = detainer_prepare_date < as.Date("2025-01-20"))

detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Encounters_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

encounter_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date),
         pre_trump = event_date < as.Date("2025-01-20"))

encounter_ids <- unique(encounter_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Admin_Arrests_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(apprehension_date = as.Date(apprehension_date),
         apprehension_method = fct(apprehension_method),
         criminality = apprehension_criminality == "1 Convicted Criminal",
         n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier"),
         pre_trump = apprehension_date < as.Date("2025-01-20")) %>% 
  filter(is.na(unique_identifier) | n_dupe == 1)

arrest_ids <- unique(arrest_df$unique_identifier)

```

```{r setup_detentions, message=FALSE, warning=FALSE, include=TRUE}

file <- "ice_detentions_nov23-feb25.csv.gz"

detain_unique_stays_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names() %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         pre_trump = stay_book_in_date_time < as.Date("2025-01-20")) %>% 
  left_join(facilities_df, by=c("detention_facility_code" = "detloc"))

```

```{r}
# Break down datasets by geography and pre vs. post Trump
segment_df <- function(df, unique_fn, geo_filter) {
  df_uq <- unique_fn(df)
  df_wa_uq <- unique_fn(geo_filter(df))
  df_pre_trump_uq <- unique_fn(filter(df, pre_trump == TRUE))
  df_post_trump_uq <- unique_fn(filter(df, pre_trump == FALSE))
  df_pre_trump_wa_uq <- unique_fn(geo_filter(filter(df, pre_trump == TRUE)))
  df_post_trump_wa_uq <- unique_fn(geo_filter(filter(df, pre_trump == FALSE)))
  
  return (list("uq"=df_uq,
            "wa_uq"=df_wa_uq,
            "pre_trump_uq"=df_pre_trump_uq,
            "post_trump_uq"=df_post_trump_uq,
            "pre_trump_wa_uq"=df_pre_trump_wa_uq,
            "post_trump_wa_uq"=df_post_trump_wa_uq))
}

detention_dfs <- segment_df(detain_unique_stays_df,
                           (function(df) return(df %>% filter(last_placement == TRUE))),
                           (function(df) return(df %>% filter(state == "WA"))))

encounter_dfs <- segment_df(encounter_df,
                           (function(df) return(df %>% arrange(desc(event_date)) %>% distinct(unique_identifier, .keep_all=TRUE))),
                           (function(df) return(df %>% filter(responsible_aor == "Seattle Area of Responsibility"))))

arrest_dfs <- segment_df(arrest_df,
                        (function(df) return(df %>% arrange(desc(apprehension_date)) %>% distinct(unique_identifier, .keep_all=TRUE))),
                        (function(df) return(df %>% filter(apprehension_aor == "Seattle Area of Responsibility"))))
```


```{r}
# Convert decimal to percentage
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}
```

This notebook attempts analyzes various "enforcement pathways" in the Pacific Northwest by joining data across encounter, arrest, and detention datasets. The following subsections analyze enforcement pathways for detentions in Washington, arrests in Seattle AOR, encounters in Seattle AOR, and NWDC detentions.
```{r}
summarize_pathways <- function(df1, df2, df3, name1, name2, name3) {
  joined_df <- df1 %>% 
    left_join(df2,
            by="unique_identifier",
            suffix=c("", paste(".", name2, sep="")),
            na_matches='never') %>% 
    left_join(df3,
              by="unique_identifier",
              suffix=c("", paste(".", name3, sep="")),
              na_matches='never')
  
  # Encounter Site, Det Facility Code
  joined_df_count <- joined_df %>%
    count(responsible_site, detention_facility_code, name="count", sort=TRUE) %>%
    mutate(pct = percent(count / nrow(joined_df)))
  print(joined_df_count)
  
  # Encounter Type, Arrest Method
  joined_df_count <- joined_df %>%
    count(event_type, apprehension_method, name="count", sort=TRUE) %>%
    mutate(pct = percent(count / nrow(joined_df)))
  print(joined_df_count)
  
  # Arrest Method, Det Facility Code
  joined_df_count <- joined_df %>% 
      count(apprehension_method, detention_facility_code, name="count", sort=TRUE) %>% 
      mutate(pct = percent(count / nrow(joined_df)))
  print(joined_df_count)
  
  # print(paste("Num Records: ", nrow(joined_df), ""))
}
```

# WA Detentions: Pre-Trump vs. Post-Trump
```{r}
summarize_pathways(detention_dfs[["pre_trump_wa_uq"]], encounter_dfs[["uq"]], arrest_dfs[["uq"]],
                   "detention", "encounter", "arrest")
summarize_pathways(detention_dfs[["post_trump_wa_uq"]], encounter_dfs[["uq"]], arrest_dfs[["uq"]],
                   "detention", "encounter", "arrest")
```

# Seattle AOR Arrests: Pre-Trump vs. Post-Trump
```{r}
summarize_pathways(arrest_dfs[["pre_trump_wa_uq"]], encounter_dfs[["uq"]], detention_dfs[["uq"]],
                   "arrest", "encounter", "detention")
summarize_pathways(arrest_dfs[["post_trump_wa_uq"]], encounter_dfs[["uq"]], detention_dfs[["uq"]],
                   "arrest", "encounter", "detention")
```

# Seattle AOR Encounters: Pre-Trump vs. Post-Trump
```{r}
summarize_pathways(encounter_dfs[["pre_trump_wa_uq"]], arrest_dfs[["uq"]], detention_dfs[["uq"]],
                   "encounter", "arrest", "detention")
summarize_pathways(encounter_dfs[["post_trump_wa_uq"]], arrest_dfs[["uq"]], detention_dfs[["uq"]],
                   "encounter", "arrest", "detention")
```

# CSCNWWA
```{r}
summarize_pathways(detention_dfs[["pre_trump_wa_uq"]] %>%
                     filter(detention_facility_code == "CSCNWWA"),
                   encounter_dfs[["uq"]], arrest_dfs[["uq"]],
                   "detention", "encounter", "arrest")
summarize_pathways(detention_dfs[["post_trump_wa_uq"]] %>%
                     filter(detention_facility_code == "CSCNWWA"),
                   encounter_dfs[["uq"]], arrest_dfs[["uq"]],
                   "detention", "encounter", "arrest")
```