---
title: "ICE ERO-LESA data Nov. 2023-Feb. 2025, Guantanamo detentions"
author: "UWCHR"
date: "2025-03-25"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

This notebook analyzes ICE enforcement data associated with detentions at Guantanamo Bay, Cuba. Datasets were obtained by the Deportation Data Project.

```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, digest, ggsankeyfier)

file <- "ice_detentions_nov23-feb25.csv.gz"

df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names() %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         )

gitmo_codes <- c("GTMOACU", "GTMODCU")

# Get all placements at Guantanamo Bay detention facilities
gitmo_detentions <- df[df$detention_facility_code %in% c("GTMOACU", "GTMODCU"),]

# Confirm one record per individual unique identifier
stopifnot(nrow(gitmo_detentions) == length(unique(gitmo_detentions$unique_identifier)))

# Get all detention history records for individuals transferred to Guantanamo Bay
gitmo_detention_histories <- df[df$unique_identifier %in% unique(gitmo_detentions$unique_identifier),]

```

ICE detention data includes `r nrow(df)` records of detention placements involving `r length(unique(df$unique_identifier))` unique individuals, identified by the field `df$unique_identifier`.

ICE detention data represents records of detention transfers or book-outs from `r min(df$detention_book_out_date_time, na.rm=TRUE)` to `r max(df$detention_book_out_date_time, na.rm=TRUE)`; therefore, some individual detention histories may be incomplete.

The dataset includes `r nrow(gitmo_detentions)` records associated with transfers to two detention locations at Guantanamo Bay, Cuba: `detention_facility_code` values  "GTMOACU" and "GTMODCU", identified as `r unique(df[df$detention_facility_code == gitmo_codes[1], "detention_facility"])` and `r unique(df[df$detention_facility_code == gitmo_codes[2], "detention_facility"])`, respectively. These detention facilities do not appear in prior detention history datasets consulted by UWCHR.

In this case, each transfer to Guantanamo Bay is associated with a single individual as indicated by the `unique_identifier` field. By filtering the dataset for all records with matching `unique_identifier` values, we find a total of `r nrow(gitmo_detention_histories)` detention history records associated with individuals transferred to Guantanamo Bay

At the time of production of the dataset, `r sum(is.na(gitmo_detentions$stay_release_reason))` placements at Guantanamo Bay facilities were ongoing, as indicated by empty `stay_release_reason` field; `r sum(!is.na(gitmo_detentions$stay_release_reason))` placement at Guantanamo Bay was complete, with `stay_release_reason` indicating `r gitmo_detentions[!is.na(gitmo_detentions$stay_release_reason), 'stay_release_reason']`.

## Transfers to Guantanamo Bay

```{r gitmo_transfers}

transfer_dates <- gitmo_detentions %>% 
  count(date(book_in_date_time))

transfer_sites <- gitmo_detentions %>% 
  count(date(book_in_date_time), detention_facility)

transfer_sites_sum <- transfer_sites %>%
  group_by(detention_facility) %>%
  summarize(n = sum(n))

transfers_from <- gitmo_detentions %>% 
  count(date(book_in_date_time), prev_facil)

```

The data record tranfers to Guantanamo Bay on `r nrow(transfer_dates)` days from `r min(transfer_dates$date)` to `r min(transfer_dates$date)`, with each transfers involving `r min(transfer_dates$n)` to `r max(transfer_dates$n)` people. It is not clear how these transfer dates compare with those documented via other reports.

Detained people transfered to Guantanamo Bay were associated with two different detention sites, with a total of `transfer_sites_sum[transfer_sites_sum$detention_facility == "JTF CAMP SIX", "n"]` transfers to "JTF CAMP SIX" and `transfer_sites_sum[transfer_sites_sum$detention_facility == "MIGRANT OPS CENTER MAIN AV622", "n"]` to "MIGRANT OPS CENTER MAIN AV622". It is not immediately apparent what is indicated by this distinction.

```{r plot_transfers}

p1 <- gitmo_detentions %>% 
  mutate(transfer_date = date(book_in_date_time)) %>% 
  count(transfer_date, detention_facility) %>% 
  ggplot(aes(x = transfer_date, y = n, fill = detention_facility)) +
  geom_col()

p1

```

## Characteristics of detained people transferred to Guantanamo Bay

```{r indiv_characteristics}

gitmo_ind <- gitmo_detentions %>% 
  distinct(unique_identifier, .keep_all = TRUE)

stopifnot(sum(gitmo_ind$gender == "Male") == nrow(gitmo_ind))

```

Recent detention history data includes a limited number of fields involving individual characteristics of detained people. (Note that in this case, each transfer to Guantanamo Bay involves a single person and there are no repeat transfers, so we could naively process the subset of transfer records to represent individual characteristics; for the sake of clarity, we generate a separate subset of records per distinct `unique_identifier` for this section.)

All individuals transferred were categorized as `gender` "Male"; where not missing, `ethnicity` is recorded as "Hispanic Origin". The only other demographic field included is `marital_status`, with the following breakdown:

```{r marital_status}

p1 <- gitmo_ind %>% 
  count(marital_status) %>% 
  ggplot(aes(y = marital_status, x = n)) +
  geom_col()

p1

```

The dataset includes a limited number of fields that relate to the legal status of 

```{r entry_status}

p1 <- gitmo_ind %>% 
  count(entry_status) %>% 
  ggplot(aes(y = entry_status, x = n)) +
  geom_col()

p1

```

Most individuals transferred to Guantanamo did not have criminal convictions.

```{r criminality}

p1 <- gitmo_ind %>% 
  count(book_in_criminality) %>% 
  ggplot(aes(y = book_in_criminality, x = n)) +
  geom_col()

p1

```

One individual was transferred to Guantanamo Bay despite lacking a final order of removal, with `case_category` encoded as "[8B] Excludable / Inadmissible - Under Adjudication by IJ", indicating ongoing immigration hearings. This invidual was the only one represented in the dataset to be transferred away from Guantanamo Bay after a placement of approximately `r gitmo_detentions[!is.na(gitmo_detentions$stay_release_reason), "placement_length"]` days.

```{r case_category}

t1 <- gitmo_ind %>% 
  count(case_category)

knitr::kable(t1)

```

As noted above, all detention stays associated with transfers to Guantanamo Bay were ongoing at the time of production of the dataset on February 18, 2025; however, we can calculate the elapsed detention stay time per individual, with a range of `r min(gitmo_detentions$stay_length_min)` to `r max(gitmo_detentions$stay_length_min)` days and an average of `r mean(gitmo_detentions$stay_length_min)` days.

```{r stay_length}

hist(gitmo_ind$stay_length_min)

```

```{r prior_stays}

# Stays associated with transfers to Guantanamo
gitmo_stays <- unique(gitmo_detentions$stayid)

# Prior completed stays for individuals associated with transfers to Guantanamo
prior_stays <- gitmo_detention_histories %>% 
  filter(!stayid %in% gitmo_stays)

prior_stays %>% 
  count(case_category)

prior_stays %>% 
  count(stay_release_reason)

```

- Join arrests characteristics
- Detention flow

```{r setup_arrests}

file <- "ERO_Admin_Arrests_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

arr <- read_excel(here('analyze', 'input', file), skip=6) %>%
  janitor::clean_names() %>%
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier")) |>
  filter(is.na(unique_identifier) | n_dupe == 1) |>
  mutate(week = floor_date(apprehension_date, "week", week_start = "Monday"))

to_join <- arr %>%
  dplyr::select(unique_identifier, citizenship_country, final_program_group)

to_join <- to_join[!duplicated(to_join$unique_identifier), ]

arr_ids <- arr$unique_identifier

df <- df %>% left_join(to_join, by='unique_identifier')

df <- df %>% 
  mutate(id_in_arr = unique_identifier %in% arr_ids)

rm(to_join)

```


```{r gtmo}

gtmo <- df[df$detention_facility_code %in% c("GTMOACU", "GTMODCU"),]

sum(gtmo$id_in_arr)

gtmo_arr <- arr[arr$unique_identifier %in% gtmo$unique_identifier,]

gtmo_arr_crim <- gtmo_arr %>% 
  count(apprehension_criminality)

gtmo_ice_arr <- gtmo[gtmo$unique_identifier %in% gtmo_arr$unique_identifier,]

gtmo_prev_facil <- gtmo %>% 
  count(prev_facil)

gtmo_first_facil <- gtmo %>% 
  count(first_facil)

gtmo_longest_facil <- gtmo %>% 
  count(longest_placement_facil)

# No records of people previously detained at NWDC--is this wrong? Or did Jose transfer to Guantanamo happen after release of data to DDP? Or not detained at NWDC?

cscnwwa_ids <- unlist(df[df$detention_facility_code == "CSCNWWA", "unique_identifier"])

sum(cscnwwa_ids %in% gtmo$unique_identifier)

```

```{r detention_flow_gitmo}

gtmo_stays <- df[df$detention_facility_code %in% c("GTMOACU", "GTMODCU"), 'stayid']

dat <- df %>% 
  filter(stayid %in% unlist(gtmo_stays)) %>% 
  dplyr::select(stayid, detention_facility_code, placement_count, stay_placements, current_placement) %>% 
  mutate(relative_placement = placement_count - stay_placements)

dat_wide <- dat %>% 
  pivot_wider(id_cols = c(stayid), names_from = relative_placement, names_prefix = "facil_n", values_from = detention_facility_code)

flows_gtmo <- dat_wide %>% 
  dplyr::select(-c(stayid)) %>% 
  group_by_all() %>% 
  summarize(n = n()) %>% 
  arrange(desc(n))

# Not sure how data should be managed to analyze flows between facilities;
# was looking at `sankeyifier`, could revisit various graph analysis packages

flows_gtmo_long <- flows_gtmo %>% 
  pivot_stages_longer(stages_from = c("facil_n-10",
"facil_n-9",
"facil_n-8",
"facil_n-7",
"facil_n-6",
"facil_n-5",
"facil_n-4",
"facil_n-3",
"facil_n-2",
"facil_n-1",
                                      "facil_n0"),
                      values_from = "n") %>% 
  filter(!is.na(node))

p1 <- ggplot(flows_gtmo_long,
       aes(x = stage, y = n, group = node,
           connector = connector, edge_id = edge_id)) +
  geom_sankeyedge(v_space = "auto", alpha=.5, aes(fill = node)) +
  geom_sankeynode(v_space = "auto") +
  labs(title = "Detention flows thru Gitmo")

pos <- position_sankey(v_space = "auto")
pos_text <- position_sankey(v_space = "auto", nudge_x = 0.1)

p1 + geom_sankeyedge(aes(fill = node), position = pos) +
  geom_sankeynode(position = pos) +
  geom_text(aes(label = node), stat = "sankeynode", position = pos_text, hjust = 0, cex = 2) +
  scale_x_discrete(expand = expansion(add = c(0.2, .6))) +
  guides(fill = FALSE) +
  theme_minimal()
```