---
title: "Unique Identifier Analysis"
author: "UWCHR"
date: "2025-06-07"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram)

```

```{r setup_facilities, message=FALSE, warning=FALSE, include=TRUE}

# Running into problem with .gz version of facilities dataset
file <- "facilities.csv"

facilities_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names()

```

```{r setup_detainers, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Detainers_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detainer_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units='days'),
         pre_trump = detainer_prepare_date < as.Date("2025-01-20"))

detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Encounters_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

encounter_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date),
         pre_trump = event_date < as.Date("2025-01-20"))

encounter_ids <- unique(encounter_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "ERO_Admin_Arrests_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(apprehension_date = as.Date(apprehension_date),
         apprehension_method = fct(apprehension_method),
         criminality = apprehension_criminality == "1 Convicted Criminal",
         n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier"),
         pre_trump = apprehension_date < as.Date("2025-01-20")) %>% 
  filter(is.na(unique_identifier) | n_dupe == 1)

arrest_ids <- unique(arrest_df$unique_identifier)

```

```{r setup_detentions, message=FALSE, warning=FALSE, include=TRUE}

file <- "ice_detentions_nov23-feb25.csv.gz"

detain_unique_stays_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names() %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         pre_trump = stay_book_in_date_time < as.Date("2025-01-20")) %>% 
  left_join(facilities_df, by=c("detention_facility_code" = "detloc"))

```

```{r}
# Break down datasets by geography and pre vs. post Trump
segment_df <- function(df, geo_filter) {
  df_wa <- geo_filter(df)
  df_pre_trump <- filter(df, pre_trump == TRUE)
  df_post_trump <- filter(df, pre_trump == FALSE)
  df_pre_trump_wa <- geo_filter(filter(df, pre_trump == TRUE))
  df_post_trump_wa <- geo_filter(filter(df, pre_trump == FALSE))
  
  return (list("wa"=df_wa,
            "pre_trump"=df_pre_trump,
            "post_trump"=df_post_trump,
            "pre_trump_wa"=df_pre_trump_wa,
            "post_trump_wa"=df_post_trump_wa))
}

detainer_dfs <- segment_df(detainer_df,
                           (function(df) return(filter(df, facility_state == "WASHINGTON"))))

detention_dfs <- segment_df(detention_df,
                            (function(df) return(filter(df, state == "WA"))))

encounter_dfs <- segment_df(encounter_df,
                            (function(df) return(filter(df, responsible_aor == "Seattle Area of Responsibility"))))

arrest_dfs <- segment_df(arrest_df,
                         (function(df) return(filter(df, apprehension_aor == "Seattle Area of Responsibility"))))
```

```{r}
# Convert decimal to percentage
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}
```

```{r}
summarize_id_counts <- function(df, print=FALSE) {
  id_count <- df %>% 
    count(unique_identifier, sort=TRUE, name="count") %>% 
    mutate(pct = percent(count / nrow(df)))
  
  count_analysis <- id_count %>% 
    count(count, sort=TRUE, name="num_ids") %>% 
    mutate(pct = percent(num_ids / nrow(id_count)))
  
  if (print) {
    print(id_count)
    print(count_analysis)
  }
  
  if (sum(is.na(id_count$unique_identifier)) == 0) {
    num_missing <- 0
    pct_missing <- 0
  } else {
    num_missing <- unlist(id_count[is.na(id_count$unique_identifier), 2], use.names=FALSE)
    pct_missing <- unlist(id_count[is.na(id_count$unique_identifier), 3], use.names=FALSE)
  }
  num_once <- unlist(count_analysis[count_analysis$count == 1, 2], use.names=FALSE)
  pct_once <- unlist(count_analysis[count_analysis$count == 1, 3], use.names=FALSE)
  id_summary <- data.frame(num_missing=num_missing,
                           pct_missing=pct_missing,
                           num_once=num_once,
                           pct_once=pct_once)
  return(id_summary)
}
```

The following tables display information about the frequency of unique identifiers in each dataset. For each section, the first table displays the number of times each identifier appears in each dataset ('count') and the proportion of identifiers that it makes up ('prop'). The second table shows the number and proportion of unique identifiers that appear 'count' number of times ('num_ids' and 'prop').

# Detainers
```{r}
detainer_id_summary <- summarize_id_counts(detainer_df)
detainer_id_summary_wa <- summarize_id_counts(detainer_dfs[["wa"]])
detainer_id_summary_pre_trump <- summarize_id_counts(detainer_dfs[["pre_trump"]])
detainer_id_summary_post_trump <- summarize_id_counts(detainer_dfs[["post_trump"]])
```

# Encounters
```{r}
encounter_id_summary <- summarize_id_counts(encounter_df)
encounter_id_summary_wa <- summarize_id_counts(encounter_dfs[["wa"]])
encounter_id_summary_pre_trump <- summarize_id_counts(encounter_dfs[["pre_trump"]])
encounter_id_summary_post_trump <- summarize_id_counts(encounter_dfs[["post_trump"]])
```

# Arrests
```{r}
arrest_id_summary <- summarize_id_counts(arrest_df)
arrest_id_summary_wa <- summarize_id_counts(arrest_dfs[["wa"]])
arrest_id_summary_pre_trump <- summarize_id_counts(arrest_dfs[["pre_trump"]])
arrest_id_summary_post_trump <- summarize_id_counts(arrest_dfs[["post_trump"]])
```

# Detention
```{r}
detention_id_summary <- summarize_id_counts(detention_df)
detention_id_summary_wa <- summarize_id_counts(detention_dfs[["wa"]])
detention_id_summary_pre_trump <- summarize_id_counts(detention_dfs[["pre_trump"]])
detention_id_summary_post_trump <- summarize_id_counts(detention_dfs[["post_trump"]])
```

# Summary of Datasets
This table summarizes key information about the frequency of unique identifiers for each datasets. For each dataset, the table displays the number and proportion of missing identifiers ('num_missing' and 'prop_missing'), and the number and proportion of unique identifiers that appear once ('num_once' and 'prop_once').

```{r}
df_name <- c("Detainer Pre-Trump", "Detainer Post-Trump",
             "Encounter Pre-Trump", "Encounter Post-Trump",
             "Arrest Pre-Trump", "Arrest Post-Trump",
             "Detention Post-Trump", "Detention Post-Trump")
all_id_summaries <- rbind(detainer_id_summary_pre_trump, detainer_id_summary_post_trump,
                          encounter_id_summary_pre_trump, encounter_id_summary_post_trump,
                          arrest_id_summary_pre_trump, arrest_id_summary_post_trump,
                          detention_id_summary_pre_trump, detention_id_summary_post_trump)
# df_name <- c("Detainer", "Detainer WA",
#              "Encounter", "Encounter WA",
#              "Arrest", "Arrest WA",
#              "Detention", "Detention WA")
# all_id_summaries <- rbind(detainer_id_summary, detainer_id_summary_wa,
#                           encounter_id_summary, encounter_id_summary_wa,
#                           arrest_id_summary, arrest_id_summary_wa,
#                           detention_id_summary, detention_id_summary_wa)
all_id_summaries <- cbind(df_name, all_id_summaries)
all_id_summaries
```