---
title: "Analysis of Detainers of Interest"
author: "UWCHR"
date: "2025-03-31"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

# Setup
```{r setup, message=FALSE, warning=FALSE, include=TRUE}
options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram)
```

```{r setup_facilities, message=FALSE, warning=FALSE, include=TRUE}
file <- "facilities.csv.gz"

facilities_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names()
```

```{r setup_detainers, message=FALSE, warning=FALSE, include=TRUE}
file <- "ERO_Detainers_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detainer_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units='days')) %>% 
  rename_with(~ paste(., "detainer", sep="."), .cols=-unique_identifier)

detainer_ids <- unique(detainer_df$unique_identifier)
```

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}
file <- "ERO_Encounters_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

encounter_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date)) %>% 
  rename_with(~ paste(., "encounter", sep="."), .cols=-unique_identifier)

encounter_ids <- unique(encounter_df$unique_identifier)
```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}
file <- "ERO_Admin_Arrests_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier")) |>
  filter(is.na(unique_identifier) | n_dupe == 1) %>%
  rename_with(~ paste(., "arrest", sep="."), .cols=-unique_identifier)

arrest_ids <- unique(arrest_df$unique_identifier)
```

```{r setup_detention, message=FALSE, warning=FALSE, include=TRUE}
file <- "ICE_Detentions_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detention_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(stay_book_in_date_time = as.Date(stay_book_in_date_time, format = "%m/%d/%Y %H:%M"),
         book_in_date_time = as.Date(book_in_date_time, format = "%m/%d/%Y %H:%M"),
         stay_book_out_date_time = as.Date(stay_book_out_date_time, format = "%m/%d/%Y %H:%M"),
         stay_book_in_date_time = as.Date(stay_book_out_date, format = "%m/%d/%Y %H:%M"),
         book_in_fy = substr(quarter(book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
         stay_book_in_fy = substr(quarter(stay_book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4)) %>% 
  left_join(facilities_df, by=c("detention_facility_code" = "detloc")) %>% 
  rename_with(~ paste(., "detention", sep="."), .cols=-unique_identifier)

detention_ids <- unique(detention_df$unique_identifier)
```


# Unique Identifier Analysis
```{r}
# Counts the number of times each unique identifier appears and the number of unique identifiers per count
summarize_id_counts <- function(df) {
  id_count <- df %>% 
    count(unique_identifier, sort=TRUE, name="count") %>% 
    mutate(prop = count / nrow(df))
  
  count_analysis <- id_count %>% 
    count(count, sort=TRUE, name="num_ids") %>% 
    mutate(prop = num_ids / nrow(id_count))
  
  print(id_count)
  print(count_analysis)
  
  num_missing <- unlist(id_count[is.na(id_count$unique_identifier), 2], use.names=FALSE)
  prop_missing <- unlist(id_count[is.na(id_count$unique_identifier), 3], use.names=FALSE)
  num_once <- unlist(count_analysis[count_analysis$count == 1, 2], use.names=FALSE)
  prop_once <- unlist(count_analysis[count_analysis$count == 1, 3], use.names=FALSE)
  id_summary <- data.frame(num_missing=num_missing,
                           prop_missing=prop_missing,
                           num_once=num_once,
                           prop_once=prop_once)
  return(id_summary)
}
```

## By Individual Dataset
The following tables display information about the frequency of unique identifiers in each dataset. For each section, the first table displays the number of times each identifier appears in each dataset ('count') and the proportion of identifiers that it makes up ('prop'). The second table shows the number and proportion of unique identifiers that appear 'count' number of times ('num_ids' and 'prop').

### Detainers
```{r}
detainer_id_summary <- summarize_id_counts(detainer_df)
```

### Encounters
```{r}
encounter_id_summary <- summarize_id_counts(encounter_df)
```

### Arrests
```{r}
arrest_id_summary <- summarize_id_counts(arrest_df)
```

### Detention
```{r}
detention_id_summary <- summarize_id_counts(detention_df)
```

## Summary of Datasets
This table summarizes key information about the frequency of unique identifiers for each datasets. For each dataset, the table displays the number and proportion of missing identifiers ('num_missing' and 'prop_missing'), and the number and proportion of unique identifiers that appear once ('num_once' and 'prop_once').
```{r}
df_name <- c("Detainer", "Encounter", "Arrest", "Detention")
all_id_summaries <- rbind(detainer_id_summary, encounter_id_summary, arrest_id_summary, detention_id_summary)
all_id_summaries <- cbind(df_name, all_id_summaries)
all_id_summaries
```

#Detainer Analysis
## Detainer/Arrest Analysis
```{r}
time_filter <- 2  # date_diff.detainer_arrest threshold (in days)

detainer_df_wa <- detainer_df %>% 
  filter(facility_state.detainer == "WASHINGTON")

detainer_arrest_df_wa <- inner_join(detainer_df_wa, arrest_df,
                                 by="unique_identifier",
                                 na_matches='never',
                                 relationship='many-to-many') %>% 
  mutate(date_diff.detainer_arrest = difftime(apprehension_date.arrest, detainer_prepare_date.detainer, units="days")) %>%
  filter(date_diff.detainer_arrest >= 0) %>%
  filter(!(detention_facility.detainer %in% c("SEATAC FED.DET.CENTER", "STATE PRISON, SHELTON", "US MARSHALS, YAKIMA", "US MARSHALS, SPOKANE,WA", "WASHINGTON STATE CORRECTIONS", "STAFFORD CREEK CORRECTIONS"))) %>%
  mutate(possible_kww_violation = date_diff.detainer_arrest <= time_filter)
```

The following charts show the frequency and dates of detainers sharing unique identifiers with arrests, without filtering by time.
```{r}
# No time filter
ggplot(detainer_arrest_df_wa, aes(x=detention_facility.detainer)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1) +
  coord_cartesian(ylim = c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Number of Detainers Sharing Unique Identifiers with Arrests, by Facility (No Time Filter)")

ggplot(detainer_arrest_df_wa, aes(x=detainer_prepare_date.detainer)) +
  geom_dotplot() +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Date Prepared of Detainers Sharing Unique Identifiers with Arrests (No Time Filter)")
```

The following charts show the frequency and dates of ``detainers of interest,'' or detainers sharing unique identifiers with arrests and placed within `r time_filter` days of an arrest. 
```{r}
# With time filter
detainers_of_interest <- detainer_arrest_df_wa %>% 
  filter(possible_kww_violation == TRUE)

ggplot(detainers_of_interest, aes(x=detention_facility.detainer)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1) +
  coord_cartesian(ylim = c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Number of Detainers Sharing Unique Identifiers with Arrests by Facility (Time Filter)")

ggplot(detainers_of_interest, aes(x=detainer_prepare_date.detainer)) +
  geom_dotplot() +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Date Prepared of Detainers Sharing Unique Identifiers with Arrests (Time Filter)")
```

In total, there are `r nrow(detainers_of_interest)` detainers of interest, as shown in the following table.
```{r}
detainers_of_interest[c("unique_identifier", "detainer_prepare_date.detainer", "apprehension_date.arrest", "date_diff.detainer_arrest", "detention_facility.detainer")]
```

The following table displays the average time between detainers and arrests with shared unique identifiers by detention facility.
```{r}
# Average time per facility
date_diff_by_facility <- detainer_arrest_df_wa %>% 
  group_by(detention_facility.detainer) %>% 
  summarize(mean=mean(date_diff.detainer_arrest))
date_diff_by_facility
```

## Enforcement Pathways
This section attempts to analyze various "enforcement pathways" in the Pacific Northwest by joining data across encounter, arrest, and detention datasets. The following subsections analyze enforcement pathways for detentions in Washington, arrests in Seattle AOR, and encounters in Seattle AOR.

```{r}

# PN suggestion: start new notebook from this section

# PN comment: Selecting distinct detention IDs will only represent a given detained person's first (or last, depending on sorting) detention record, is this what we want? (Same if multuple arrests/encounters/detainers but less of an issue because fewer repeat IDs)

detention_df_wa_uq <- detention_df %>% 
  filter(state.detention == "WA") %>% 
  arrange(desc(stay_book_in_date_time.detention)) %>% 
  distinct(unique_identifier, .keep_all=TRUE)

detention_df_uq <- detention_df %>% 
  arrange(desc(stay_book_in_date_time.detention)) %>% 
  distinct(unique_identifier, .keep_all=TRUE)

encounter_df_wa_uq <- encounter_df %>% 
  filter(responsible_aor.encounter == "Seattle Area of Responsibility") %>% 
  arrange(desc(event_date.encounter)) %>% 
  distinct(unique_identifier, .keep_all=TRUE)

encounter_df_uq <- encounter_df %>% 
  arrange(desc(event_date.encounter)) %>% 
  distinct(unique_identifier, .keep_all=TRUE)

arrest_df_wa_uq <- arrest_df %>% 
  filter(apprehension_aor.arrest == "Seattle Area of Responsibility") %>% 
  arrange(desc(apprehension_date.arrest)) %>% 
  distinct(unique_identifier, .keep_all=TRUE)

arrest_df_uq <- arrest_df %>% 
  arrange(desc(apprehension_date.arrest)) %>% 
  distinct(unique_identifier, .keep_all=TRUE)

```

```{r}
summarize_pathways <- function(df1, df2, df3) {
  joined_df <- df1 %>% 
    left_join(df2,
            by="unique_identifier",
            na_matches='never') %>% 
    left_join(df3,
              by="unique_identifier",
              na_matches='never')
  
  joined_df_count_geo <- joined_df %>% 
    count(responsible_site.encounter, detention_facility_code.detention, name="count", sort=TRUE)
  print(joined_df_count_geo)
  
  joined_df_count_method <- joined_df %>% 
    count(event_type.encounter, apprehension_method.arrest, name="count", sort=TRUE)
  print(joined_df_count_method)
  
  print(nrow(joined_df))
}
```

## Detention
```{r}
summarize_pathways(detention_df_wa_uq, encounter_df_uq, arrest_df_uq)
```

## Arrest
```{r}
summarize_pathways(arrest_df_wa_uq, encounter_df_uq, detention_df_uq)
```

## Encounter
```{r}
summarize_pathways(encounter_df_wa_uq, arrest_df_uq, detention_df_uq)
```

#PN thoughts on next steps:

- Look at difference in these enforcement pathways post-Trump inauguration, compared to earlier period of dataset and/or last three months of Biden admin. Changes in rank or relationship between encounter-arrest-detention locations, methods?
- Look at full detention pathways, not just initial book-in (double check what is happening when you select a single record per unique ID in detentions data). Changes post-Trump? Separate analysis for detentions not in arrests dataset, or not in WA. (Detentions with no associated arrest either represent transfers from CBP or, less likely, arrests pre-dating release of dataset.)
- Any other changes pre-post Trump? Rate of missingness of IDs or other data, for example?