---
title: "Analysis of Detainers of Interest"
author: "UWCHR"
date: "2025-03-31"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

# Setup
```{r setup, message=FALSE, warning=FALSE, include=TRUE}
options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram)
```

```{r setup_facilities, message=FALSE, warning=FALSE, include=TRUE}
file <- "facilities.csv"

facilities_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  janitor::clean_names()
```

```{r setup_detainers, message=FALSE, warning=FALSE, include=TRUE}
file <- "ERO_Detainers_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detainer_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units='days'),
         pre_trump = detainer_prepare_date < as.Date("2025-01-20"))

detainer_ids <- unique(detainer_df$unique_identifier)
```

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}
file <- "ERO_Encounters_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

encounter_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date),
         pre_trump = event_date < as.Date("2025-01-20"))

encounter_ids <- unique(encounter_df$unique_identifier)
```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}
file <- "ERO_Admin_Arrests_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(apprehension_date = as.Date(apprehension_date),
         apprehension_method = fct(apprehension_method),
         criminality = apprehension_criminality == "1 Convicted Criminal",
         n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier"),
         pre_trump = apprehension_date < as.Date("2025-01-20")) %>% 
  filter(is.na(unique_identifier) | n_dupe == 1)

arrest_ids <- unique(arrest_df$unique_identifier)
```

```{r setup_detention, message=FALSE, warning=FALSE, include=TRUE}
file <- "ICE_Detentions_2025-ICLI-00019_2024-ICFO-39357_LESA-STU_FINAL_raw.xlsx"

detention_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(stay_book_in_date_time = as.Date(stay_book_in_date_time, format = "%m/%d/%Y %H:%M"),
         book_in_date_time = as.Date(book_in_date_time, format = "%m/%d/%Y %H:%M"),
         stay_book_out_date_time = as.Date(stay_book_out_date_time, format = "%m/%d/%Y %H:%M"),
         stay_book_in_date_time = as.Date(stay_book_out_date, format = "%m/%d/%Y %H:%M"),
         book_in_fy = substr(quarter(book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
         stay_book_in_fy = substr(quarter(stay_book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
         pre_trump = book_in_date_time < as.Date("2025-01-20")) %>% 
  left_join(facilities_df, by=c("detention_facility_code" = "detloc"))

detention_ids <- unique(detention_df$unique_identifier)
```

```{r}
# Break down datasets by geography and pre vs. post Trump
segment_df <- function(df, geo_filter) {
  df_wa <- geo_filter(df)
  df_pre_trump <- filter(df, pre_trump == TRUE)
  df_post_trump <- filter(df, pre_trump == FALSE)
  df_pre_trump_wa <- geo_filter(filter(df, pre_trump == TRUE))
  df_post_trump_wa <- geo_filter(filter(df, pre_trump == FALSE))
  
  return (list("wa"=df_wa,
            "pre_trump"=df_pre_trump,
            "post_trump"=df_post_trump,
            "pre_trump_wa"=df_pre_trump_wa,
            "post_trump_wa"=df_post_trump_wa))
}

detainer_dfs <- segment_df(detainer_df,
                           (function(df) return(filter(df, facility_state == "WASHINGTON"))))

detention_dfs <- segment_df(detention_df,
                            (function(df) return(filter(df, state == "WA"))))

encounter_dfs <- segment_df(encounter_df,
                            (function(df) return(filter(df, responsible_aor == "Seattle Area of Responsibility"))))

arrest_dfs <- segment_df(arrest_df,
                         (function(df) return(filter(df, apprehension_aor == "Seattle Area of Responsibility"))))
```


```{r}
# Convert decimal to percentage
percent <- function(x, digits = 2, format = "f", ...) {
  paste0(formatC(100 * x, format = format, digits = digits, ...), "%")
}
```


# Unique Identifier Analysis

```{r}
summarize_id_counts <- function(df, print=FALSE) {
  id_count <- df %>% 
    count(unique_identifier, sort=TRUE, name="count") %>% 
    mutate(pct = percent(count / nrow(df)))
  
  count_analysis <- id_count %>% 
    count(count, sort=TRUE, name="num_ids") %>% 
    mutate(pct = percent(num_ids / nrow(id_count)))
  
  if (print) {
    print(id_count)
    print(count_analysis)
  }
  
  if (sum(is.na(id_count$unique_identifier)) == 0) {
    num_missing <- 0
    pct_missing <- 0
  } else {
    num_missing <- unlist(id_count[is.na(id_count$unique_identifier), 2], use.names=FALSE)
    pct_missing <- unlist(id_count[is.na(id_count$unique_identifier), 3], use.names=FALSE)
  }
  num_once <- unlist(count_analysis[count_analysis$count == 1, 2], use.names=FALSE)
  pct_once <- unlist(count_analysis[count_analysis$count == 1, 3], use.names=FALSE)
  id_summary <- data.frame(num_missing=num_missing,
                           pct_missing=pct_missing,
                           num_once=num_once,
                           pct_once=pct_once)
  return(id_summary)
}
```

## By Individual Dataset
The following tables display information about the frequency of unique identifiers in each dataset. For each section, the first table displays the number of times each identifier appears in each dataset ('count') and the proportion of identifiers that it makes up ('prop'). The second table shows the number and proportion of unique identifiers that appear 'count' number of times ('num_ids' and 'prop').

### Detainers
```{r}
detainer_id_summary <- summarize_id_counts(detainer_df)
detainer_id_summary_wa <- summarize_id_counts(detainer_dfs[["wa"]])
detainer_id_summary_pre_trump <- summarize_id_counts(detainer_dfs[["pre_trump"]])
detainer_id_summary_post_trump <- summarize_id_counts(detainer_dfs[["post_trump"]])
```

### Encounters
```{r}
encounter_id_summary <- summarize_id_counts(encounter_df)
encounter_id_summary_wa <- summarize_id_counts(encounter_dfs[["wa"]])
encounter_id_summary_pre_trump <- summarize_id_counts(encounter_dfs[["pre_trump"]])
encounter_id_summary_post_trump <- summarize_id_counts(encounter_dfs[["post_trump"]])
```

### Arrests
```{r}
arrest_id_summary <- summarize_id_counts(arrest_df)
arrest_id_summary_wa <- summarize_id_counts(arrest_dfs[["wa"]])
arrest_id_summary_pre_trump <- summarize_id_counts(arrest_dfs[["pre_trump"]])
arrest_id_summary_post_trump <- summarize_id_counts(arrest_dfs[["post_trump"]])
```

### Detention
```{r}
detention_id_summary <- summarize_id_counts(detention_df)
detention_id_summary_wa <- summarize_id_counts(detention_dfs[["wa"]])
detention_id_summary_pre_trump <- summarize_id_counts(detention_dfs[["pre_trump"]])
detention_id_summary_post_trump <- summarize_id_counts(detention_dfs[["post_trump"]])
```

## Summary of Datasets
This table summarizes key information about the frequency of unique identifiers for each datasets. For each dataset, the table displays the number and proportion of missing identifiers ('num_missing' and 'prop_missing'), and the number and proportion of unique identifiers that appear once ('num_once' and 'prop_once').

```{r}
df_name <- c("Detainer Pre-Trump", "Detainer Post-Trump",
             "Encounter Pre-Trump", "Encounter Post-Trump",
             "Arrest Pre-Trump", "Arrest Post-Trump",
             "Detention Post-Trump", "Detention Post-Trump")
all_id_summaries <- rbind(detainer_id_summary_pre_trump, detainer_id_summary_post_trump,
                          encounter_id_summary_pre_trump, encounter_id_summary_post_trump,
                          arrest_id_summary_pre_trump, arrest_id_summary_post_trump,
                          detention_id_summary_pre_trump, detention_id_summary_post_trump)
# df_name <- c("Detainer", "Detainer WA",
#              "Encounter", "Encounter WA",
#              "Arrest", "Arrest WA",
#              "Detention", "Detention WA")
# all_id_summaries <- rbind(detainer_id_summary, detainer_id_summary_wa,
#                           encounter_id_summary, encounter_id_summary_wa,
#                           arrest_id_summary, arrest_id_summary_wa,
#                           detention_id_summary, detention_id_summary_wa)
all_id_summaries <- cbind(df_name, all_id_summaries)
all_id_summaries
```

#Detainer Analysis
## Detainer/Arrest Analysis
```{r}
time_filter <- 2  # date_diff.detainer_arrest threshold (in days)

detainer_arrest_wa_df <- inner_join(detainer_dfs[["wa"]], arrest_df,
                                    by="unique_identifier",
                                    suffix=c(".detainer", ".arrest"),
                                    na_matches='never',
                                    relationship='many-to-many') %>% 
  mutate(date_diff.detainer_arrest = difftime(apprehension_date.arrest, detainer_prepare_date, units="days")) %>%
  filter(date_diff.detainer_arrest >= 0) %>%
  filter(!(detention_facility %in% c("SEATAC FED.DET.CENTER", "STATE PRISON, SHELTON", "US MARSHALS, YAKIMA", "US MARSHALS, SPOKANE,WA", "WASHINGTON STATE CORRECTIONS", "STAFFORD CREEK CORRECTIONS"))) %>%
  mutate(possible_kww_violation = date_diff.detainer_arrest <= time_filter)
```


The following charts show the frequency and dates of detainers sharing unique identifiers with arrests, without filtering by time.
```{r}
# No time filter
ggplot(detainer_arrest_df_wa, aes(x=detention_facility.detainer)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1) +
  coord_cartesian(ylim = c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Number of Detainers Sharing Unique Identifiers with Arrests, by Facility (No Time Filter)")

ggplot(detainer_arrest_df_wa, aes(x=detainer_prepare_date.detainer)) +
  geom_dotplot() +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Date Prepared of Detainers Sharing Unique Identifiers with Arrests (No Time Filter)")
```

The following charts show the frequency and dates of ``detainers of interest,'' or detainers sharing unique identifiers with arrests and placed within `r time_filter` days of an arrest. 
```{r}
# With time filter
detainers_of_interest <- detainer_arrest_df_wa %>% 
  filter(possible_kww_violation == TRUE)

ggplot(detainers_of_interest, aes(x=detention_facility.detainer)) +
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1) +
  coord_cartesian(ylim = c(0, 15)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ggtitle("Number of Detainers Sharing Unique Identifiers with Arrests by Facility (Time Filter)")

ggplot(detainers_of_interest, aes(x=detainer_prepare_date.detainer)) +
  geom_dotplot() +
  theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("Date Prepared of Detainers Sharing Unique Identifiers with Arrests (Time Filter)")
```

In total, there are `r nrow(detainers_of_interest)` detainers of interest, as shown in the following table.
```{r}
detainers_of_interest[c("unique_identifier", "detainer_prepare_date.detainer", "apprehension_date.arrest", "date_diff.detainer_arrest", "detention_facility.detainer")]
```

The following table displays the average time between detainers and arrests with shared unique identifiers by detention facility.
```{r}
# Average time per facility
date_diff_by_facility <- detainer_arrest_df_wa %>% 
  group_by(detention_facility.detainer) %>% 
  summarize(mean=mean(date_diff.detainer_arrest))
date_diff_by_facility
```