---
title: "ICE ERO-LESA detention data Oct. 2022 - Mar. 2026"
author: "UWCHR"
date: "2026-04-01"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---


```{r setup, message=FALSE, warning=FALSE, include=FALSE}

options(scipen = 1000000)
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
options(ggplot2.discrete.colour = "viridis")
options(ggplot2.discrete.fill = "viridis")

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, digest, ggsankeyfier)

every_nth = function(n) {
  return(function(x) {x[c(TRUE, rep(FALSE, n - 1))]})
}

every_2nd_nth = function(n) {
  return(function(x) {x[c(FALSE, rep(TRUE, n - 1))]})
}

arrests <- arrow::read_parquet("https://github.com/deportationdata/ice/raw/refs/heads/main/data/arrests-latest.parquet")

detentions <- arrow::read_parquet("https://github.com/deportationdata/ice/raw/refs/heads/main/data/detention-stints-latest.parquet")

headcount <- arrow::read_parquet("https://github.com/deportationdata/ice/raw/refs/heads/main/data/facilities-daily-population-latest.parquet")

facilities <- arrow::read_parquet("https://github.com/deportationdata/ice-detention-facilities/raw/refs/heads/main/data/facilities-latest.parquet")

detention_stays <- read_delim(here::here("detain-unique-stays", "output", "detention-stays-nov22-mar26.csv.gz"), delim='|')

```

# Analysis

(NA transfers/release reason mean still detained at NWIPC as of March 2026)

```{r transfers_from_cscnwwa, message=FALSE, warning=FALSE, echo=FALSE}

cscnwwa_detentions <- detentions %>% 
  filter(detention_facility_code == "CSCNWWA")

cscnwwa_transfers <- detentions %>% 
  filter(detention_facility_code == "CSCNWWA",
         detention_release_reason == "Transferred")

cscnwwa_transfers_by_gender <- cscnwwa_transfers %>%
  count(next_facil, gender)

p1 <- cscnwwa_transfers_by_gender %>%
  filter(gender != "Unknown") %>%
  ggplot(aes(x = gender, y = n, fill = next_facil)) +
  geom_col(position="fill") +
  labs(title = "Transfer destinations from NWIPC by gender")

ggplotly(p1)

cscnwwa_bookouts_by_gender <- cscnwwa_detentions %>% 
  filter(gender != "Unknown") %>%
  mutate(detention_release_reason = case_when(str_detect(detention_release_reason, "Paroled") ~ "Paroled",
                                              str_detect(detention_release_reason, "Order of Supervision") ~ "Order of supervision",
                                              str_detect(detention_release_reason, "Order of Recognizance") ~ "Order of recognizance",
                                              str_detect(detention_release_reason, "ORR") ~ "ORR",
                                              str_detect(detention_release_reason, "Bonded Out") ~ "Bonded out",
                                              TRUE ~ detention_release_reason)) %>% 
  count(detention_release_reason, gender)

p1 <- cscnwwa_bookouts_by_gender %>% 
  filter(gender != "Unknown") %>% 
  ggplot(aes(x = gender, y = n, fill = detention_release_reason)) +
  geom_col(position="fill") +
  labs(title ="NWIPC detention stint release reason by age, gender")

ggplotly(p1)

```

```{r}

cscnwwa_transfers_by_gender_age <- cscnwwa_transfers %>% 
  mutate(age = year(book_in_date_time) - birth_year,
         age_group = cut(age,
                           breaks = c(0, 18, 25, 40, 60, Inf),
                           right = FALSE)) %>% 
  count(next_facil, gender, age_group)

p1 <- cscnwwa_transfers_by_gender_age %>% 
    filter(gender != "Unknown",
         age_group != "[0,18)") %>% 
  ggplot(aes(x = gender, y = n, fill = next_facil)) +
  geom_col(position="fill") +
  facet_wrap(~age_group) +
  labs(title ="NWIPC transfer facility by age, gender")

ggplotly(p1)


cscnwwa_bookouts_by_gender_age <- cscnwwa_detentions %>% 
  mutate(age = year(book_in_date_time) - birth_year,
         age_group = cut(age,
                           breaks = c(0, 18, 25, 40, 60, Inf),
                           right = FALSE),
         detention_release_reason = case_when(str_detect(detention_release_reason, "Paroled") ~ "Paroled",
                                              str_detect(detention_release_reason, "Order of Supervision") ~ "Order of supervision",
                                              str_detect(detention_release_reason, "Order of Recognizance") ~ "Order of recognizance",
                                              str_detect(detention_release_reason, "ORR") ~ "ORR",
                                              str_detect(detention_release_reason, "Bonded Out") ~ "Bonded out",
                                              TRUE ~ detention_release_reason)) %>% 
  count(detention_release_reason, gender, age_group)

p1 <- cscnwwa_bookouts_by_gender_age %>% 
  filter(gender != "Unknown",
         age_group != "[0,18)") %>% 
  ggplot(aes(x = gender, y = n, fill = detention_release_reason)) +
  geom_col(position="fill") +
  facet_wrap(~age_group) +
  labs(title ="NWIPC detention stint release reason by age, gender")

ggplotly(p1)

```

```{r}

dat <- detention_stays %>% 
  filter(gender != "Unknown",
         !is.na(book_in_date_time)) %>% 
  mutate(age = year(book_in_date_time) - birth_year,
         age_group = cut(age,
                           breaks = c(0, 18, 25, 40, 60, Inf),
                           right = FALSE),
         age_gender_group = paste(age_group, gender))

p1 <- dat %>% 
  ggplot(aes(y = age_gender_group, x = total_placements)) +
  geom_boxplot()

p1

```

```{r}

p1 <- dat %>% 
  filter(book_in_aor == "Miami Area of Responsibility") %>% 
  ggplot(aes(y = age_gender_group, x = total_placements)) +
  geom_boxplot() +
  facet_wrap(~book_in_aor)

p1

p1 <- dat %>% 
  filter(book_in_aor == "Seattle Area of Responsibility") %>% 
  ggplot(aes(y = age_gender_group, x = total_placements)) +
  geom_boxplot() +
  facet_wrap(~book_in_aor)

p1

```

```{r}

dat <- detention_stays %>% 
  filter(gender != "Unknown",
         !is.na(book_in_date_time)) %>% 
  mutate(age = year(book_in_date_time) - birth_year,
         age_group = cut(age,
                           breaks = c(0, 18, 25, 40, 60, Inf),
                           right = FALSE),
         age_gender_group = paste(age_group, gender),
         det_since_2025 = year(stay_book_in_date_time) >= 2025,
         stay_book_in_year = year(stay_book_in_date_time))

p1 <- dat %>% 
  filter(stay_book_in_year > 2021) %>% 
  ggplot(aes(y = age_gender_group, x = total_placements)) +
  geom_boxplot() +
  facet_wrap(~stay_book_in_year)

p1

p2 <- dat %>% 
  filter(stay_book_in_year > 2021) %>% 
  ggplot(aes(y = as.factor(stay_book_in_year), x = total_placements)) +
  geom_boxplot()

p2

p3 <- dat %>% 
  filter(stay_book_in_year > 2021,
         !is.na(age_group)) %>% 
  ggplot(aes(y = as.factor(stay_book_in_year), x = total_placements)) +
  geom_boxplot() +
  facet_wrap(age_group~gender)

p3


p3 <- dat %>% 
  filter(stay_book_in_year > 2021,
         !is.na(age_group)) %>% 
  ggplot(aes(y = age_gender_group, x = total_placements)) +
  geom_boxplot() +
  facet_wrap(~det_since_2025) +
  labs(title = "Total detention placements by age/gender",
       subtitle = "TRUE = detention stays starting since Jan. 1, 2025")

p3
```


Would it be possible to see whether of women <40 booked into the Miami AOR in November 2025, the number of transfers looks unusual (compared to men booked into that region during that month; compared to overall people booked into that region during that month; compared to women <40 booked into other facilities nationwide in November 2025; compared to all people booked into other facilities nationwide in November 2025)? 

```{r}

nov_2025_bookins <- detentions %>% 
  mutate(age = year(book_in_date_time) - birth_year,
         age_group = cut(age,
                           breaks = c(0, 18, 25, 40, 60, Inf),
                           right = FALSE),
         age_gender_group = paste(age_group, gender)) %>% 
  filter(as.yearmon(stay_book_in_date_time) == as.yearmon("2025-11-01"),
         gender != "Unknown",
         !is.na(age),
         age >= 18) %>% 
  mutate(group = "natl")

nov_2025_bookins_mia <- nov_2025_bookins %>% 
  filter(book_in_aor == "Miami Area of Responsibility") %>% 
  mutate(group = "mia")

nov_2025_bookins_distinct <- nov_2025_bookins %>% 
  distinct(stay_ID, .keep_all = TRUE)

nov_2025_bookins_mia_distinct <- nov_2025_bookins_mia %>% 
  distinct(stay_ID, .keep_all = TRUE)

dat <- rbind(nov_2025_bookins_distinct, nov_2025_bookins_mia_distinct)

p1 <- dat %>% 
  ggplot(aes(x = total_placements, y = age_gender_group)) +
  geom_boxplot() +
  facet_wrap(~group) +
  labs(title = "Detention stays starting in Nov. 2025, total stints by age/gender",
       subtitle = "Women under 40 booked into Miami AOR experienced more transfers")

p1

```

chart over time of the average number of transfers experienced by all people booked into Elizabeth Contract Detention Facility, Delaney Hall, or Moshannon Valley Processing Center?

```{r}

facil_list <- c('ELZICDF', 'DHDFNJ', 'MSVPCPA')

detention_stays <- detentions %>% 
  distinct(stay_ID, .keep_all = TRUE)

dat_facil <- detention_stays %>% 
  filter(stay_book_in_date_time >= "2022-10-01" &
           stay_book_in_date_time <= "2025-12-31",
         detention_facility_code %in% facil_list) %>% 
  mutate(stay_book_in_month = as.yearmon(stay_book_in_date_time)) %>% 
  group_by(detention_facility, detention_facility_code, stay_book_in_month) %>% 
  summarize(avg_placements = mean(total_placements),
            med_placements = median(total_placements)) %>% 
  mutate(group = detention_facility) %>% 
  ungroup() %>% 
  dplyr::select(-c(detention_facility, detention_facility_code))

dat_natl <- detention_stays %>% 
  filter(stay_book_in_date_time >= "2022-10-01" &
           stay_book_in_date_time <= "2025-12-31") %>% 
  mutate(stay_book_in_month = as.yearmon(stay_book_in_date_time)) %>% 
  group_by(stay_book_in_month) %>% 
  summarize(avg_placements = mean(total_placements),
            med_placements = median(total_placements)) %>% 
  mutate(group = "NATIONWIDE")

dat <- rbind(dat_natl, dat_facil)

p1 <- dat %>%
  ggplot(aes(x = stay_book_in_month, y = avg_placements, color = group)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Avg. number of detention stints by stay book-in month",
       subtitle = "Selected facilities versus nationwide")

p1

```

```{r}

dat_natl <- detention_stays %>% 
  filter(stay_book_in_date_time >= "2022-10-01" &
           stay_book_in_date_time <= "2025-12-31") %>% 
  mutate(stay_book_in_month = as.yearmon(stay_book_in_date_time)) %>% 
  group_by(stay_book_in_month) %>% 
  summarize(avg_placements = mean(total_placements),
            med_placements = median(total_placements),
            avg_n_facil = mean(n_facil),
            med_n_facil = median(n_facil)) %>% 
  mutate(group = "NATIONWIDE")

dat_aor <- detention_stays %>% 
  filter(stay_book_in_date_time >= "2022-10-01" &
           stay_book_in_date_time <= "2025-12-31",
         book_in_aor == "Miami Area of Responsibility") %>% 
  mutate(stay_book_in_month = as.yearmon(stay_book_in_date_time)) %>% 
  group_by(stay_book_in_month) %>% 
  summarize(avg_placements = mean(total_placements),
            med_placements = median(total_placements),
            avg_n_facil = mean(n_facil),
            med_n_facil = median(n_facil)) %>% 
  mutate(group = "MIA")

dat <- rbind(dat_natl, dat_aor)

p1 <- dat %>% 
  ggplot(aes(x = stay_book_in_month, y = avg_n_facil, color = group, group=group)) +
  geom_line() +
  ylim(0, NA) +
  theme_minimal() +
  labs(title = "Average number of unique facilities per stay",
       subtitle = "By stay book-in month and AOR (Miami versus nationwide)")

p1

p2 <- dat %>% 
  ggplot(aes(x = stay_book_in_month, y = avg_placements, color = group, group=group)) +
  geom_line() +
  ylim(0, NA) +
  theme_minimal() +
  labs(title = "Average number of detention stints per stay",
       subtitle = "By stay book-in month and AOR (Miami versus nationwide)")

p2

write_delim(dat, "transfer_data.csv", delim=",")

```

```{r}

b1 <- detention_stays %>% 
  filter(stay_book_in_date_time >= "2022-10-01" &
           stay_book_in_date_time <= "2025-12-31") %>% 
  mutate(stay_book_in_month = as.yearmon(stay_book_in_date_time)) %>% 
  ggplot(aes(x = stay_book_in_month, y = n_facil, group=stay_book_in_month)) +
  geom_boxplot()

b1

```

# Hold rooms

```{r}

detentions <- detentions %>% 
  mutate(stint_length = difftime(book_out_date_time, book_in_date_time, units = "days"),
         stint_over_72 = stint_length > 3,
         holdroom = str_detect(detention_facility_code, "HOLD"))

det_stint_summary <- detentions %>% 
  group_by(detention_facility_code) %>% 
  summarize(avg_stint = mean(stint_length, na.rm=TRUE),
            med_stint = median(stint_length, na.rm=TRUE)) %>% 
  ungroup()

med_under_72 <- unlist(det_stint_summary[det_stint_summary$med_stint < 3, "detention_facility_code"])

detentions <- detentions %>% 
  mutate(med_stint_under_72 = detention_facility_code %in% med_under_72)

p1 <- detentions %>% 
  filter(book_in_date_time >= "2022-10-01") %>% 
  filter(holdroom == TRUE) %>%
  mutate(book_in_month = as.yearmon(book_in_date_time)) %>% 
  ggplot(aes(x = book_in_month, y = log(as.numeric(stint_length)), group=book_in_month)) +
  geom_boxplot() +
  geom_hline(yintercept = log(3), color="red") +
  labs(title = "Distribution of ICE hold room stint lengths by book-in month")

p1

```

```{r}

dat_natl <- detentions %>% 
  filter(book_in_date_time >= "2022-10-01",
         holdroom == TRUE) %>% 
  mutate(book_in_month = as.yearmon(book_in_date_time)) %>% 
  group_by(book_in_month) %>% 
  summarize(n = n(),
            avg_stint = mean(stint_length, na.rm=TRUE),
            med_stint = median(stint_length, na.rm=TRUE),
            n_stint_over_72 = sum(stint_over_72, na.rm=TRUE),
            pct_stint_over_72 = n_stint_over_72 / n * 100
            ) %>% 
  mutate(group = "NATIONWIDE")

write_delim(dat_natl, "ice_hold_room_summary.csv", delim=',')

p1 <- dat_natl %>% 
  ggplot(aes(x = book_in_month, y = n_stint_over_72)) +
  geom_line() +
  theme_minimal() +
  labs(title = "ICE hold room detention stints longer than 72 hours",
       subtitle = "By detention stint book-in month")

p1

```

```{r}

dat_per_facil <- detentions %>% 
  filter(book_in_date_time >= "2022-10-01",
         holdroom == TRUE) %>% 
  mutate(book_in_month = as.yearmon(book_in_date_time)) %>% 
  group_by(book_in_month, detention_facility_code) %>% 
  summarize(n = n(),
            avg_stint = mean(stint_length, na.rm=TRUE),
            med_stint = median(stint_length, na.rm=TRUE),
            n_stint_over_72 = sum(stint_over_72, na.rm=TRUE),
            pct_stint_over_72 = n_stint_over_72 / n * 100
            ) %>% 
  mutate(group = detention_facility_code) %>% 
  dplyr::select(-detention_facility_code)

p1 <- dat_per_facil %>% 
  filter(group == "IWAHOLD") %>% 
  ggplot(aes(x = book_in_month, y = n_stint_over_72)) +
  geom_line() +
  theme_minimal() +
  labs(title = "AROCC detention stints longer than 72 hours",
       subtitle = "By detention stint book-in month")

p1

```

```{r}

dat <- rbind(dat_natl, dat_per_facil)

p1 <- dat %>% 
  filter(group %in% c("NATIONWIDE", "IWAHOLD")) %>% 
  ggplot(aes(x = book_in_month, y = n_stint_over_72, color = group)) +
  geom_line() +
  theme_minimal() +
  labs(title = "ICE hold room detention stints longer than 72 hours",
       subtitle = "By detention stint book-in month")

p1
  

```

```{r}

p1 <- detention_stays %>% 
  ggplot(aes(y = first_facil_aor, x = n_aors)) +
  geom_boxplot() +
  scale_x_continuous(breaks=seq(1,10)) +
  labs(title = "Distribution of detainee transfers",
       subtitle = "total distinct AORs by book-in facility AOR")

p1

dat_natl <- detention_stays %>% 
  filter(book_in_date_time >= "2022-10-01") %>% 
  mutate(book_in_month = as.yearmon(book_in_date_time)) %>% 
  group_by(book_in_month) %>% 
  summarize(avg_aors = mean(n_aors, na.rm=TRUE),
            med_aors = median(n_aors, na.rm=TRUE)) %>% 
  ungroup() %>% 
  mutate(group = "NATIONWIDE")

dat_new <- detention_stays %>% 
  filter(book_in_date_time >= "2022-10-01",
         first_facil_aor == "Newark") %>% 
  mutate(book_in_month = as.yearmon(book_in_date_time)) %>% 
  group_by(book_in_month, first_facil_aor) %>% 
  summarize(avg_aors = mean(n_aors, na.rm=TRUE),
            med_aors = median(n_aors, na.rm=TRUE)) %>% 
  ungroup() %>% 
  dplyr::select(-first_facil_aor) %>% 
  mutate(group = "NEW")

dat <- rbind(dat_natl, dat_new)

p2 <- dat %>% 
  ggplot(aes(x = book_in_month, y = avg_aors, color = group, group=group)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Detainees experience more transfers between AORs",
       subtitle = "By detention stay book-in month, nationwide versus Newark AOR")

p2
  

```
