---
title: "Analysis of removals dataset"
author: "UWCHR"
date: "2025-07-11"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---


```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram, UpSetR)

```

- Removals from Seattle AOR
- Removals of ppl arrested in WA
- Removals by whether detained at NWDC
- Ppl transferred from NWDC to SEAHOLD: patterns
- Ppl arrested in WA and removed w/o being held at NWDC
- Ppl removed via methods excluding ICE Air hubs

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

# file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Encounters.xlsx"
# 
# encounter_df_1 <- read_excel(here('analyze', 'input', file), sheet=1, skip=6) %>%
#   janitor::clean_names() %>%
#   mutate(event_date = as.Date(event_date))
# 
# encounter_df_2 <- read_excel(here('analyze', 'input', file), sheet=2, skip=6) %>%
#   janitor::clean_names() %>%
#   mutate(event_date = as.Date(event_date))
# 
# encounter_df <- rbind(encounter_df_1, encounter_df_2)
# 
# rm(encounter_df_1, encounter_df_2)
# 
# encounter_ids <- unique(encounter_df$unique_identifier)

```


```{r setup_detainers}

# file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Detainers.xlsx"
# 
# detainer_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
#   janitor::clean_names()
# 
# detainer_df<- detainer_df%>% 
#   mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units = "days"))
# 
# detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Admin_Arrests_LESA-STU_FINAL_Redacted_raw.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names()

arrest_df <- arrest_df %>% 
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("unique_identifier")) |>
  filter(!is.na(unique_identifier))

arrest_ids <- unique(arrest_df$unique_identifier)

arrests_with_removal <- arrest_df %>% 
  filter(!is.na(departed_date))

sea_aor_arrests <- arrest_df %>% 
  filter(apprehension_aor == "Seattle Area of Responsibility")

wa_arrests <- arrest_df %>% 
  filter(apprehension_state == "WASHINGTON")

```

```{r setup_detention, message=FALSE, warning=FALSE, include=TRUE}

file <- "ice_detentions_nov23-jul25.csv.gz"

detention_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         stay_book_out_date = ymd(stay_book_out_date),
         book_in_fy = substr(quarter(book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
         stay_book_in_fy = substr(quarter(stay_book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4)) %>% filter(!is.na(unique_identifier))


detention_df <- detention_df %>% 
  group_by(unique_identifier) %>% 
  arrange(stay_book_in_date_time, book_in_date_time, unique_identifier) %>% 
  ungroup()

detention_ids <- unique(detention_df$unique_identifier)

detention_removals <- detention_df %>% 
  filter(stay_release_reason == "Removed")

cscnwwa_detentions <- detention_df %>% 
  filter(detention_facility_code == "CSCNWWA")

wa_facil <- c("FDLHOLD",
"HAVMCWA",
"RICHOLD",
"SPOHOLD",
"STJMCWA",
"TACGHWA",
"WNTHOLD",
"ABRDNWA",
"ADAMSWA",
"BLHHOLD",
"BENTOWA",
"CHELAWA",
"CLJUVWA",
"CLARKWA",
"COWLIWA",
"COWJVWA",
"ECOLOWA",
"FERRYWA",
"FORKSWA",
"FRANKWA",
"GRANTWA",
"KINCOWA",
"KITTIWA",
"MARTIWA",
"MASONWA",
"WAMCNEI",
"CSCNWWA",
"OKANOWA",
"PENDOWA",
"RRFINWA",
"REGJCWA",
"BOPSET",
"SEAICDF",
"SEAHOLD",
"SPOKAWA",
"STEVEWA",
"SUNNYWA",
"THURSWA",
"WHICOWA",
"YIKIMWA",
"YAKHOLD")

wa_detentions <- detention_df %>% 
  filter(detention_facility_code %in% wa_facil)

```

```{r setup_removals, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ICE_Removals07312025.xlsx"

removal_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% filter(!is.na(unique_identifier))

removal_ids <- unique(removal_df$unique_identifier)

removal_df <- removal_df %>% 
  mutate(wa_arrest = unique_identifier %in% wa_arrests$unique_identifier,
         sea_aor_arrest = unique_identifier %in% sea_aor_arrests$unique_identifier,
         nwdc_detention = unique_identifier %in% cscnwwa_detentions$unique_identifier,
         wa_detention = unique_identifier %in% wa_detentions$unique_identifier)

```

```{r compare_sets}

data_levels <- c("arrest", "detention", "removal")

sets <- list("arrests" = unique(arrests_with_removal$unique_identifier),
             "detentions" = unique(detention_removals$unique_identifier),
             "removals" = removal_ids)

# p1 <- ggVennDiagram(x = sets,
#                     color = name,
#                     label = "count") +
#   scale_fill_gradient(low = "#F4FAFE", high = "#4981BF")
# 
# p1

p2 <- ggVennDiagram(x = sets,
              force_upset = TRUE, order.set.by = "name", order.intersect.by = "size"
              )

p2 

```

Strong weekly periodicity

```{r}

dat <- removal_df %>% 
  count(departed_date)

p1 <- dat %>% 
  ggplot(aes(x = departed_date, y = n)) +
  geom_line()
  
p1

```

```{r}

dat <- removal_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week)

p1 <- dat %>% 
  ggplot(aes(x = week, y = n)) +
  geom_col()
  
p1

```

Note HQ AOR important for removals

```{r}

dat <- removal_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(week, docket_aor)

p1 <- dat %>% 
  ggplot(aes(x = week, y = n)) +
  geom_col() +
  facet_wrap(~docket_aor)
  
p1

```

Most removals Tues-Fri

```{r}

dat <- removal_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         weekday = wday(departed_date, week_start = "Monday", label=TRUE, abbr=TRUE),
         weeknum = week(departed_date),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(weekday)

p1 <- dat %>% 
  ggplot(aes(x = weekday, y = n)) +
  geom_col()
  
p1

```

Removals by weekday and week in 2025. Not very interesting

```{r}

dat <- removal_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         weekday = wday(departed_date, week_start = "Monday", label=TRUE, abbr=TRUE),
         weeknum = week(departed_date),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(weekday, weeknum)

p1 <- dat %>% 
  ggplot(aes(x = weekday, y = n, color = weeknum)) +
  geom_line(aes(group=weeknum))
  
p1

```

```{r docket_aor}

tab <- removal_df %>% 
  count(docket_aor) %>% 
  arrange(desc(n))

```

```{r sea_removals}

p0 <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week,) %>% 
  ggplot(aes(x = week, y = n)) +
  geom_col()

p0

p1 <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, sea_aor_arrest) %>% 
  ggplot(aes(x = week, y = n, fill = sea_aor_arrest)) +
  geom_col()

p1

p2 <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, wa_arrest) %>% 
  ggplot(aes(x = week, y = n, fill = wa_arrest)) +
  geom_col()

p2

p3 <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, nwdc_detention) %>% 
  ggplot(aes(x = week, y = n, fill = nwdc_detention)) +
  geom_col()

p3

```

```{r tab_docket}

tab_docket <- removal_df %>%
  filter(departed_date >= "2025-01-01") %>% 
  count(docket_aor)

tab_docket_wa <- removal_df %>%
  filter(departed_date >= "2025-01-01") %>% 
  count(docket_aor, wa_arrest)

tab_docket_nwdc <- removal_df %>%
  filter(departed_date >= "2025-01-01") %>% 
  count(docket_aor, nwdc_detention)

```

```{r}

dat <- removal_df %>% 
  filter(wa_arrest == TRUE) %>% 
  count(docket_aor)

p1 <- dat %>% 
  mutate(docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  ggplot(aes(y = docket_aor, x = n)) +
  geom_col() +
  labs(title = "WA ICE arrests by removal docket AOR")

p1

```

```{r docket_aor_port_of_depart}

tab_sea_pod <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  count(port_of_departure)

tab_sea_pod_2025 <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility",
         departed_date >= "2025-01-20") %>% 
  count(port_of_departure) %>% 
  arrange(desc(n))

tab_sea_pod_2025_india <- removal_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility",
         departed_date >= "2025-01-20") %>% 
  count(port_of_departure, citizenship_country) %>% 
  arrange(desc(n)) %>% 
  filter(citizenship_country == "INDIA")


```

```{r docket_aor_port_of_depart}

tab_sea_cit <- removal_df %>% 
  filter(departed_date >= "2025-01-01") %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  count(citizenship_country)

tab_sea_dep <- removal_df %>% 
  filter(departed_date >= "2025-01-01") %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  count(departure_country)

tab_sea_cit_dep <- removal_df %>% 
  filter(departed_date >= "2025-01-01") %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  count(citizenship_country, departure_country)

```

Note records designated as removals of Venezuelans to El Salvador are not inclusive of all Venezuelans expelled to CECOT

```{r venezuelans_to_el_salvador}

arr_ven_es <- arrest_df %>% 
  filter(!is.na(departed_date),
         !is.na(unique_identifier),
         citizenship_country == 'VENEZUELA',
         departure_country == 'EL SALVADOR')

det_ven_es <- detention_df %>% 
  filter(!is.na(departed_date),
         !is.na(unique_identifier),
         citizenship_country == 'VENEZUELA',
         departure_country == 'EL SALVADOR')

rem_ven_es <- removal_df %>% 
  filter(!is.na(departed_date),
         !is.na(unique_identifier),
         citizenship_country == 'VENEZUELA',
         departure_country == 'EL SALVADOR')

ven_es_ids <- union(union(arr_ven_es$unique_identifier, det_ven_es$unique_identifier), rem_ven_es$unique_identifier)

venezuelan_removals_el_salvador <- removal_df %>% 
  filter(unique_identifier %in% ven_es_ids)

write_delim(arr_ven_es, here::here('analyze', 'output', 'arr_ven_es.csv'), delim='|')
write_delim(det_ven_es, here::here('analyze', 'output', 'det_ven_es.csv'), delim='|')
write_delim(rem_ven_es, here::here('analyze', 'output', 'rem_ven_es.csv'), delim='|')

```

```{r}

rem_ven_cecot <- removal_df %>% 
  filter(departed_date == as.Date("2025-03-15")|
           departed_date == as.Date("2025-03-31")|
           departed_date == as.Date("2025-04-13"),
         !is.na(unique_identifier),
         citizenship_country == 'VENEZUELA')

rem_ven_apr_may <- removal_df %>% 
  filter(departed_date >= as.Date("2025-03-01") & departed_date <= as.Date("2025-04-30"),
         !is.na(unique_identifier),
         citizenship_country == "VENEZUELA")

rem_ven_apr_4 <- removal_df %>% 
  filter(departed_date == as.Date("2025-04-04"),
         !is.na(unique_identifier),
         citizenship_country == 'VENEZUELA')

```

```{r rem_ven}

dat <- removal_df %>% 
  filter(citizenship_country == "VENEZUELA")

p1 <- dat %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  count(yearmon, departure_country) %>% 
  ggplot(aes(x = yearmon, y = n, fill = departure_country)) +
  geom_col() +
  gghighlight(departure_country == "EL SALVADOR")

p1

p2 <- dat %>% 
  mutate(yearmon = as.yearmon(departed_date),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(yearmon, docket_aor) %>% 
  ggplot(aes(x = yearmon, y = n, fill = docket_aor)) +
  geom_col() +
  gghighlight(docket_aor == "Seattle")

p2

p3 <- dat %>% 
  filter(gender == "Female") %>% 
  count(departed_date, gender) %>% 
  ggplot(aes(x = departed_date, y = n, fill = gender)) +
  geom_col()

p3

p4 <- rem_ven_cecot %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  count(yearmon, gender) %>% 
  ggplot(aes(x = yearmon, y = n, fill = gender)) +
  geom_col()

p4

temp <- rem_ven_cecot %>% 
  filter(gender == "Female")

```

```{r}

cecot_flight_det_histories <- detention_df %>% 
  filter(unique_identifier %in% rem_ven_cecot$unique_identifier,
         as.Date(detention_book_out_date_time) == as.Date("2025-03-15"))

dat <- detention_df %>% 
  filter(
         citizenship_country == "VENEZUELA")  %>% 
  filter(book_in_date_time >= as.Date('2025-01-01')) %>% 
  mutate(book_in_date = as.Date(book_in_date_time))

dat_top_facil <- dat %>% 
  count(detention_facility_code) %>% 
  arrange(desc(n))

p1 <- dat %>% 
  count(book_in_date, detention_facility_code) %>% 
  ggplot(aes(x=book_in_date, y = n, fill = detention_facility_code)) + 
  geom_col() +
  gghighlight(detention_facility_code %in% c("WEBDCTX", "RGRNDTX", "ELVDFTX"))

p1

```

```{r}

cecot_flight_arr_histories <- arrest_df %>% 
  filter(unique_identifier %in% rem_ven_cecot$unique_identifier)

cecot_flight_arr_histories %>% 
  count(apprehension_aor) %>% 
  arrange(desc(n))

```

## Salvadorans

```{r rem_sv}

rem_sv <- removal_df %>% 
  filter(citizenship_country == "EL SALVADOR") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"))

tab1 <- rem_sv %>% 
  count(citizenship_country, departure_country)

write_delim(tab1, here::here('analyze', 'output', 'removals-salvadorans-departed-country.csv'))

dat <- rem_sv %>% 
  count(week, departure_country)

write_delim(dat, here::here('analyze', 'output', 'weekly-removals-salvadorans-departed_country.csv'))

p1 <- dat %>% 
  ggplot(aes(x = week, y = n, fill = departure_country)) +
  geom_col() +
  gghighlight(departure_country %in% c("EL SALVADOR", "MEXICO")) +
  labs(title = "Weekly total removals of Salvadorans by departure country",
       subtitle = "El Salvador, Mexico highlighted", 
       caption = "Source: ICE FOIA 2025-ICLI-00019_2024-ICFO-39357")
  

p1


```

```{r rem_to_sv}

rem_to_sv <- removal_df %>% 
  filter(departure_country == "EL SALVADOR") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"))

tab2 <- rem_to_sv %>% 
  count(citizenship_country, departure_country)

write_delim(tab1, here::here('analyze', 'output', 'removals-to-el-salvador-by-citizenship.csv'))

dat2 <- rem_to_sv %>% 
  count(week, citizenship_country)

write_delim(dat2, here::here('analyze', 'output', 'weekly-removals-to-el-salvador-by-citizenship.csv'))

p2 <- dat2 %>% 
  ggplot(aes(x = week, y = n, fill = citizenship_country)) +
  geom_col() +
  gghighlight(citizenship_country %in% c("EL SALVADOR", "VENEZUELA", "GUATEMALA", "HONDURAS", "NICARAGUA", "COLOMBIA"))+
  labs(title = "Weekly total removals to El Salvador by citizenship country",
       subtitle = "Select nationalities highlighted", 
       caption = "Source: ICE FOIA 2025-ICLI-00019_2024-ICFO-39357")
  
p2

```

## Comparing various datasets/subsets for removal trends over time

```{r historic_removals}

old_removals <- read_delim(here::here("analyze", "input", "ice_removals_fy12-23ytd.csv.gz"), delim = '|') %>% 
  mutate(departed_date = as.Date(departed_date, format = "%m/%d/%Y"))

old_removals <- old_removals %>% 
  mutate(month = floor_date(departed_date, "month"))

new_removals <- removal_df %>% 
  mutate(month = floor_date(departed_date, "month"),
         area_of_responsibility = docket_aor)

dat1 <- old_removals %>% 
  count(month, area_of_responsibility) %>% 
  mutate(source = "uwchr-removals")
  
dat2 <- new_removals %>% 
  count(month, area_of_responsibility) %>% 
  mutate(source = "ddp-removals")

dat <- rbind(dat1, dat2)

dat_1 <- dat %>% 
  group_by(month, source) %>% 
  summarize(n = sum(n))

p1 <- dat_1 %>% 
  ggplot(aes(x = month, y = n, fill = source)) +
  geom_col() +
  labs(title = "Total monthly removals",
       subtitle = "Nationwide")

p1

```

```{r}

dat1 <- old_removals %>% 
  count(month, area_of_responsibility) %>% 
  mutate(source = "uwchr-removals")
  
dat2 <- new_removals %>% 
  mutate(area_of_responsibility = docket_aor) %>% 
  count(month, area_of_responsibility) %>% 
  mutate(source = "ddp-removals")

dat <- rbind(dat1, dat2)

write_delim(dat, here::here('analyze', 'output', 'monthly_removals_by_aor_uwchr_ddp.csv'), delim=',')

p1 <- dat %>% 
  mutate(area_of_responsibility = str_replace(area_of_responsibility, " Area of Responsibility", "")) %>% 
  group_by(month, area_of_responsibility) %>% 
  summarize(n = sum(n)) %>% 
  ggplot(aes(x = month, y = n)) +
  geom_col() +
  facet_wrap(~area_of_responsibility)

p1

p2 <- dat %>% 
  filter(area_of_responsibility == "Seattle Area of Responsibility") %>% 
  group_by(month, area_of_responsibility, source) %>% 
  summarize(n = sum(n)) %>% 
  ggplot(aes(x = month, y = n, fill = source)) +
  geom_col() +
  labs(title = "Total monthly removals",
       subtitle = "Seattle Area of Responsibility")

p2

```

```{r historic_removals}

old_removals <- old_removals %>% 
  mutate(month = floor_date(departed_date, "month"))

new_removals <- removal_df %>% 
  mutate(month = floor_date(departed_date, "month"))

det_removals <- detention_df %>% 
  filter(detention_release_reason == "Removed") %>% 
  mutate(month = floor_date(departed_date, "month"))

arr_removals <- arrests_with_removal %>% 
  mutate(month = floor_date(departed_date, "month"))

dat1 <- old_removals %>% 
  count(month) %>% 
  mutate(source = "uwchr-removals")
  
dat2 <- new_removals %>% 
  count(month) %>% 
  mutate(source = "ddp-removals")

dat3 <- det_removals %>% 
  count(month) %>% 
  mutate(source = "ddp-detentions-with-removal")

dat4 <- arr_removals %>% 
  count(month) %>% 
  mutate(source = "ddp-arrests-with-removal")

dat <- rbind(dat1, dat2, dat3, dat4)

p1 <- dat %>% 
  filter(month >= "2011-10-01") %>% 
  group_by(month, source) %>% 
  summarize(n = sum(n)) %>% 
  ggplot(aes(x = month, y = n, fill = source)) +
  geom_col(position='dodge') +
  labs(title = "Total weekly removals, nationwide")

p1

```

