---
title: "Analysis of removals dataset"
author: "UWCHR"
date: "2025-07-11"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---


```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, plotly, gghighlight, viridis, readxl, ggVennDiagram, UpSetR)

```

- Removals from Seattle AOR
- Removals of ppl arrested in WA
- Removals by whether detained at NWDC
- Ppl transferred from NWDC to SEAHOLD: patterns
- Ppl arrested in WA and removed w/o being held at NWDC
- Ppl removed via methods excluding ICE Air hubs

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

# file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Encounters_FINAL_unprotected.xlsx"
# 
# encounter_df_1 <- read_excel(here('analyze', 'input', file), sheet=1, skip=6) %>%
#   janitor::clean_names() %>%
#   mutate(event_date = as.Date(event_date))
# 
# encounter_df_2 <- read_excel(here('analyze', 'input', file), sheet=2, skip=6) %>%
#   janitor::clean_names() %>%
#   mutate(event_date = as.Date(event_date))
# 
# encounter_df <- rbind(encounter_df_1, encounter_df_2)
# 
# rm(encounter_df_1, encounter_df_2)
# 
# encounter_ids <- unique(encounter_df$unique_identifier)

```


```{r setup_detainers}

# file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Detainers_LESA-STU_FINAL_unprotected.xlsx"
# 
# detainer_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
#   janitor::clean_names()
# 
# detainer_df<- detainer_df%>% 
#   mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units = "days"))
# 
# detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Admin_Arrests_LESA-STU_FINAL_unprotected.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=5) %>%
  janitor::clean_names()

arrest_df <- arrest_df %>%
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("unique_identifier")) |>
  filter(!is.na(unique_identifier))

arrest_ids <- unique(arrest_df$unique_identifier)

sea_aor_arrests <- arrest_df %>% 
  filter(apprehension_aor == "Seattle Area of Responsibility")

wa_arrests <- arrest_df %>% 
  filter(apprehension_state == "WASHINGTON")

```

```{r setup_detention, message=FALSE, warning=FALSE, include=TRUE}

# file <- "ice_detentions_nov23-jun25.csv.gz"
# 
# detention_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
#   mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
#          book_in_date_time = ymd_hms(book_in_date_time),
#          detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
#          stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
#          stay_book_out_date = ymd(stay_book_out_date),
#          book_in_fy = substr(quarter(book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
#          stay_book_in_fy = substr(quarter(stay_book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4)) %>% filter(!is.na(unique_identifier))
# 
# 
# detention_df <- detention_df %>% 
#   group_by(unique_identifier) %>% 
#   arrange(stay_book_in_date_time, book_in_date_time, unique_identifier) %>% 
#   ungroup()
# 
# detention_ids <- unique(detention_df$unique_identifier)

```

```{r setup_removals, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ICE_Removals_LESA-STU_FINAL_2025only_unprotected.xlsx"

removals_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names() %>% filter(!is.na(unique_identifier))

removal_ids <- unique(removals_df$unique_identifier)

removals_df <- removals_df %>% 
  mutate(wa_arrest = unique_identifier %in% wa_arrests$unique_identifier,
         sea_aor_arrest = unique_identifier %in% sea_aor_arrests$unique_identifier)

```

Strong weekly periodicity

```{r}

dat <- removals_df %>% 
  count(departed_date)

p1 <- dat %>% 
  ggplot(aes(x = departed_date, y = n)) +
  geom_col()
  
p1

```

```{r}

dat <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week)

p1 <- dat %>% 
  ggplot(aes(x = week, y = n)) +
  geom_col()
  
p1

```

Note HQ AOR important for removals

```{r}

dat <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(week, docket_aor)

p1 <- dat %>% 
  ggplot(aes(x = week, y = n)) +
  geom_col() +
  facet_wrap(~docket_aor)
  
p1

```

Most removals Tues-Fri

```{r}

dat <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         weekday = wday(departed_date, week_start = "Monday", label=TRUE, abbr=TRUE),
         weeknum = week(departed_date),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(weekday)

p1 <- dat %>% 
  ggplot(aes(x = weekday, y = n)) +
  geom_col()
  
p1

```

Removals by weekday and week in 2025. Not very interesting

```{r}

dat <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         weekday = wday(departed_date, week_start = "Monday", label=TRUE, abbr=TRUE),
         weeknum = week(departed_date),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(weekday, weeknum)

p1 <- dat %>% 
  ggplot(aes(x = weekday, y = n, color = weeknum)) +
  geom_line(aes(group=weeknum))
  
p1

```

```{r docket_aor}

tab <- removals_df %>% 
  count(docket_aor) %>% 
  arrange(desc(n))

```

```{r sea_removals}

p0 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week,) %>% 
  ggplot(aes(x = week, y = n)) +
  geom_col()

p0

p1 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, sea_aor_arrest) %>% 
  ggplot(aes(x = week, y = n, fill = sea_aor_arrest)) +
  geom_col()

p1

p2 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, wa_arrest) %>% 
  ggplot(aes(x = week, y = n, fill = wa_arrest)) +
  geom_col()

p2

```

```{r docket_aor_port_of_depart}

tab_sea_pod <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  count(port_of_departure)

tab_sea_pod_2025 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility",
         departed_date >= "2025-01-20") %>% 
  count(port_of_departure) %>% 
  arrange(desc(n))

tab_sea_pod_2025 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility",
         departed_date >= "2025-01-20") %>% 
  count(port_of_departure, citizenship_country) %>% 
  arrange(desc(n)) %>% 
  filter(citizenship_country == "INDIA")


```


