---
title: "Comparison of anonymized identifiers across datasets"
author: "UWCHR"
date: "2025-03-31"
output:
    html_document:
        html_preview: true
        toc: true
        toc_depth: 3
        toc_float: true
        code_folding: hide
---

- Third country removals
	- of people arrested in SEA
	- of people detained at CSCNWWA
	- from SEA directly

```{r setup, message=FALSE, warning=FALSE, include=TRUE}

options(scipen = 1000000)

library(pacman)

p_load(here, tidyverse, zoo, lubridate, ggplot2, ggforce, plotly, gghighlight, viridis, readxl, ggVennDiagram)

```

```{r setup_detainers, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Detainers_LESA-STU_FINAL_unprotected.xlsx"

detainer_df<- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names()

detainer_df<- detainer_df%>% 
  mutate(date_diff = difftime(detainer_prepare_date, apprehension_date, units = "days"))

detainer_ids <- unique(detainer_df$unique_identifier)

```

```{r setup_encounters, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Encounters_FINAL_unprotected.xlsx"

encounter_df_1 <- read_excel(here('analyze', 'input', file), sheet=1, skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date))

encounter_df_2 <- read_excel(here('analyze', 'input', file), sheet=2, skip=6) %>% 
  janitor::clean_names() %>% 
  mutate(event_date = as.Date(event_date))

encounter_df <- rbind(encounter_df_1, encounter_df_2)

rm(encounter_df_1, encounter_df_2)

encounter_ids <- unique(encounter_df$unique_identifier)

```

```{r setup_arrests, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ERO_Admin_Arrests_LESA-STU_FINAL_unprotected.xlsx"

arrest_df<- read_excel(here('analyze', 'input', file), skip=5) %>% 
  janitor::clean_names()

arrest_df <- arrest_df %>% 
  mutate(apprehension_date = as.Date(apprehension_date)) |>
  mutate(apprehension_method = fct(apprehension_method)) |>
  mutate(criminality = apprehension_criminality == "1 Convicted Criminal") |>
  mutate(n_dupe = row_number(), .by = c("apprehension_date", "apprehension_method", "apprehension_criminality", "unique_identifier")) |>
  filter(is.na(unique_identifier) | n_dupe == 1)

arrest_ids <- unique(arrest_df$unique_identifier)

```

```{r setup_detention, message=FALSE, warning=FALSE, include=TRUE}

file <- "ice_detentions_nov23-jun25.csv.gz"

detention_df <- read_delim(here('analyze', 'input', file), delim='|') %>% 
  mutate(stay_book_in_date_time = ymd_hms(stay_book_in_date_time),
         book_in_date_time = ymd_hms(book_in_date_time),
         detention_book_out_date_time = ymd_hms(detention_book_out_date_time),
         stay_book_out_date_time = ymd_hms(stay_book_out_date_time),
         stay_book_out_date = ymd(stay_book_out_date),
         book_in_fy = substr(quarter(book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4),
         stay_book_in_fy = substr(quarter(stay_book_in_date_time, fiscal_start = 10, type = "year.quarter"), 1, 4))


detention_df <- detention_df %>% 
  group_by(unique_identifier) %>% 
  arrange(stay_book_in_date_time, book_in_date_time, unique_identifier) %>% 
  ungroup()

detention_ids <- unique(detention_df$unique_identifier)

```

```{r setup_removals, message=FALSE, warning=FALSE, include=TRUE}

file <- "2025-ICLI-00019_2024-ICFO-39357_ICE_Removals_LESA-STU_FINAL_2025only_unprotected.xlsx"

removals_df <- read_excel(here('analyze', 'input', file), skip=6) %>% 
  janitor::clean_names()

removal_ids <- unique(removals_df$unique_identifier)

```

```{r compare_sets}

sets <- list("encounters" = encounter_ids,
             "detainers" = detainer_ids,
             "arrests" = arrest_ids,
             "detentions" = detention_ids,
             "removals" = removal_ids)

vd1 <- ggVennDiagram(x = sets,
              force_upset = TRUE, order.set.by = "name", order.intersect.by = "size"
              )

vd1 

```

```{r setup_id_intersect_cols}

detention_df <- detention_df %>%
  mutate(id_in_arrests = unique_identifier %in% arrest_ids,
         id_in_detainers = unique_identifier %in% detainer_ids,
         id_in_removals = unique_identifier %in% removal_ids)

arrest_df <- arrest_df |>
  mutate(id_in_detentions = unique_identifier %in% detention_ids,
         id_in_detainers = unique_identifier %in% detainer_ids,
         id_in_removals = unique_identifier %in% removal_ids)

detainer_df<- detainer_df |>
  mutate(id_in_arrests = unique_identifier %in% arrest_ids,
         id_in_detainers = unique_identifier %in% detainer_ids,
         id_in_removals = unique_identifier %in% removal_ids)

removals_df <- removals_df |>
  mutate(id_in_detentions = unique_identifier %in% detention_ids,
         id_in_detainers = unique_identifier %in% detainer_ids,
         id_in_arrests = unique_identifier %in% arrest_ids)

```

Only a minority of arrest and removal IDs coincide, despite arrests covering longer time period than removals.  Trend for arrests dataset "looks better" than trend for removals dataset. Note DDP signals problems with removals dataset.

```{r plot_id_time}

sets <- list("arrests" = arrest_ids,
             "removals" = removal_ids)

vd1 <- ggVennDiagram(x = sets,
              force_upset = TRUE, order.set.by = "name", order.intersect.by = "size"
              )

vd1 

p1 <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, id_in_arrests) %>% 
  ggplot(aes(x = week, y = n, color = id_in_arrests)) +
  geom_line()

p1

p2 <- arrest_df %>% 
  mutate(week = floor_date(apprehension_date, "week", week_start = "Monday")) %>% 
  count(week, id_in_removals) %>% 
  ggplot(aes(x = week, y = n, color = id_in_removals)) +
  geom_line()

p2

```

# Arrests/removals

```{r national_arrests_removals}

dat1 <- arrest_df %>% 
  mutate(week = floor_date(apprehension_date, "week", week_start = "Monday")) %>% 
  count(week) %>% 
  mutate(type = "arrests")

dat2 <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week) %>% 
  mutate(type = "removals")

dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  ggplot(aes(x=week, y=n, color=type)) +
  geom_line() +
  ylim(0, NA)

p1

```

```{r aor_arrests_removals}

dat1 <- arrest_df %>% 
  mutate(week = floor_date(apprehension_date, "week", week_start = "Monday"),
         aor = str_replace(apprehension_aor, " Area of Responsibility", "")) %>% 
  count(week, aor) %>% 
  mutate(type = "arrests")

dat2 <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(week, aor) %>% 
  mutate(type = "removals")

dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  ggplot(aes(x=week, y=n, color=type)) +
  geom_line() +
  facet_wrap(~aor)
  ylim(0, NA)

p1

p1.1 <- dat %>% 
  ggplot(aes(x=week, y=n, color=type)) +
  geom_line() +
  facet_wrap(~aor, scales = "free_y")

p1.1

```

Note `apprehension_aor` and `docket_aor` do not coincide.

```{r diff_arrests_removals}

dat_wide <- pivot_wider(dat, id_cols=c("week", "aor"), names_from = "type", values_from = "n") %>% 
  filter(week >= "2024-12-30") %>% 
  mutate(arrests_minus_removals = arrests - removals)

p2 <- dat_wide %>% 
  ggplot(aes(x=week, y=arrests_minus_removals)) +
  geom_line() +
  facet_wrap(~aor) +
  geom_link2(aes(colour = after_stat(ifelse(y > 0, "positve", "negative"))))

p2

```

Arrests dataset also has `departed_date` and `departed_country` values associated with removals. But these account for only a partial share of removals documented in removals dataset. One reason would be removals resulting from arrests which occurred prior to scope of arrests dataset. For this reason probably best to rely on removals dataset for analysis of this phenomenon?

```{r compare_removals_arrests}

dat1 <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week) %>% 
  mutate(type = "removals")

dat2 <- arrest_df %>% 
  filter(apprehension_date >= as.Date("2023-11-01"),
         departed_date >= as.Date("2023-11-01"),
          !is.na(departed_date)) %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week) %>% 
  mutate(type = "arrests with removal data")

dat <- rbind(dat1, dat2)

p1 <- dat %>% 
  ggplot(aes(x=week, y=n, color=type)) +
  geom_line()

p1

```

# Third country deportations

All datasets include `citizenship_country` and `departure_country` fields which could indicate third-country deportations. Most comprehensive set of potential cases is drawn from combination of databases (here focusing on arrests, detentions, removals). Be alert for discrepancies between these datasets.

```{r third_country_deportations_ids}

removals_third_country_ids <- removals_df %>% 
  filter(departure_country != citizenship_country) %>% 
  dplyr::select(unique_identifier)

arrests_third_country_ids <- arrest_df %>% 
  filter(!is.na(departure_country)) %>% 
  filter(departure_country != citizenship_country) %>% 
  dplyr::select(unique_identifier)

detention_third_country_ids <- detention_df %>% 
  filter(!is.na(departure_country)) %>% 
  filter(departure_country != citizenship_country) %>% 
  dplyr::select(unique_identifier)

third_country_ids <- union(removals_third_country_ids, arrests_third_country_ids)

third_country_ids <- unlist(union(third_country_ids, detention_third_country_ids))

```

```{r removals_df_only}

removals_df <- removals_df %>% 
  mutate(third_country_deport = unique_identifier %in% third_country_ids)

p1 <- removals_df %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  count(yearmon, third_country_deport) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_deport)) +
  geom_line()

p1

p1.1 <- removals_df %>% 
  mutate(yearmon = as.yearmon(departed_date),
         docket_aor = str_replace_all(docket_aor, " Area of Responsibility", "")) %>% 
  count(yearmon, docket_aor, third_country_deport) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_deport)) +
  geom_line() +
  facet_wrap(~docket_aor)

p1.1

```

```{r pct_third_country_natl}

dat <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, third_country_deport) %>% 
  mutate(type = "national")

dat_wide <- dat %>% 
  pivot_wider(id_cols = c("week"), names_from = "third_country_deport", names_prefix = "third_country_", values_from = "n") %>% 
  group_by(week) %>% 
  mutate(pct_third_country = third_country_TRUE / sum(third_country_TRUE, third_country_FALSE),
         type = "removals dataset")

p1 <- dat_wide %>% 
  ggplot(aes(x = week, y = pct_third_country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Third-country deportations as % of weekly total",
       subtitle = "Nationwide, since Jan. 2025",
       caption = "Source: ICE 'removals' dataset obtained by DDP")

p1

```

```{r pct_third_country_natl_via_arrests}

arrest_df <- arrest_df %>% 
  mutate(third_country_deport = unique_identifier %in% third_country_ids)

dat1 <- arrest_df %>% 
  filter(!is.na(departure_country),
    departed_date >= "2023-11-01") %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday")) %>% 
  count(week, third_country_deport) %>% 
  mutate(type = "arrests dataset")

dat1_wide <- dat1 %>% 
  pivot_wider(id_cols = c("week"), names_from = "third_country_deport", names_prefix = "third_country_", values_from = "n") %>% 
  group_by(week) %>% 
  mutate(pct_third_country = third_country_TRUE / sum(third_country_TRUE, third_country_FALSE),
         type = "national_via_arrests")

p1 <- dat1_wide %>% 
  ggplot(aes(x = week, y = pct_third_country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Third-country deportations as % of weekly total",
       subtitle = "Nationwide, since Nov. 2023",
       caption = "Source: ICE 'arrests' dataset obtained by DDP")

p1

```

```{r compare}

compare_dat <- rbind(dat_wide, dat1_wide)

p1 <- compare_dat %>% 
  ggplot(aes(x = week, y = pct_third_country, color = type)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Third-country deportations as % of weekly total",
       subtitle = "Comparison of records drawn from arrests versus removals datasets")

p1

```

```{r pct_third_country_aor}

dat <- removals_df %>% 
  mutate(week = floor_date(departed_date, "week", week_start = "Monday"),
         docket_aor = str_replace(docket_aor, " Area of Responsibility", "")) %>% 
  count(week, third_country_deport, docket_aor) %>% 
  mutate(type = "national")

dat_wide <- dat %>% 
  pivot_wider(id_cols = c("week", "docket_aor"), names_from = "third_country_deport", names_prefix = "third_country_", values_from = "n") %>% 
  group_by(week, docket_aor) %>% 
  mutate(pct_third_country = third_country_TRUE / sum(third_country_TRUE, third_country_FALSE))

p1 <- dat_wide %>% 
  ggplot(aes(x = week, y = pct_third_country)) +
  geom_line() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Third-country deportations as % of weekly total",
       subtitle = "per Docket AOR, since Jan. 2025") +
  facet_wrap(~docket_aor)

p1

```

```{r citizenship_departure_country}

tab1 <- removals_df %>% 
  filter(third_country_deport == TRUE) %>% 
  count(citizenship_country, departure_country) %>% 
  arrange(desc(n))

p1 <- tab1 %>% 
  mutate(name = paste(citizenship_country, departure_country, sep = "-"),
         name = fct_reorder(name, desc(n))) %>% 
  head(10) %>% 
  ggplot(aes(y = name, x = n)) +
  geom_col() +
  labs(title = "Top 10 citizenship-departure country pairs")

p1

tab1.1 <- removals_df %>% 
  filter(third_country_deport == TRUE) %>% 
  count(citizenship_country) %>% 
  mutate(citizenship_country = fct_reorder(citizenship_country, desc(n))) %>% 
  arrange(desc(n))

p1.1 <- tab1.1 %>% 
  head(10) %>% 
  ggplot(aes(y = citizenship_country, x = n)) +
  geom_col() +
  labs(title = "Top 10 third-country removals by citizenship")

p1.1  

tab1.2 <- removals_df %>% 
  filter(third_country_deport == TRUE) %>% 
  count(departure_country) %>% 
  mutate(departure_country = fct_reorder(departure_country, desc(n))) %>% 
  arrange(desc(n))

p1.2 <- tab1.2 %>% 
  head(10) %>% 
  ggplot(aes(y = departure_country, x = n)) +
  geom_col() +
  labs(title = "Top 10 third-country removals by departure country")

p1.2 

```

How to get all Seattle area third-country removals:
- All Seattle AOR arrests, removal docket, and detentions
- With third country deport characteristics

Need:
- Detention facility AOR

```{r seattle_third_country_removals}

seattle_arrest_ids <- arrest_df[arrest_df$apprehension_aor == "Seattle Area of Responsibility", "unique_identifier"]

# This is not all-inclusive, `docket_aor` is misleading filter
# Looking for records with Seattle AOR docket or arrest in Seattle AOR
seattle_third_country_removals <- removals_df %>% 
  filter(third_country_deport == TRUE,
         docket_aor == "Seattle Area of Responsibility" | unique_identifier %in% unlist(seattle_arrest_ids))

# Minimal pattern except Venezuelans to Mexico
tab2 <- seattle_third_country_removals %>% 
  count(citizenship_country, departure_country) %>% 
  arrange(desc(n))

knitr::kable(head(tab2, 10))

```


```{r arrests_df_only}

arrest_df <- arrest_df %>% 
  mutate(third_country_deport = departure_country != citizenship_country)

p1 <- arrest_df %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  filter(yearmon >= as.yearmon("2023-11-01")) %>% 
  count(yearmon, third_country_deport) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_deport)) +
  geom_line()

p1

p1.1 <- arrest_df %>% 
  mutate(yearmon = as.yearmon(departed_date),
         apprehension_aor = str_replace_all(apprehension_aor, " Area of Responsibility", "")) %>% 
  filter(yearmon >= as.yearmon("2023-11-01")) %>% 
  count(yearmon, apprehension_aor, third_country_deport) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_deport)) +
  geom_line() +
  facet_wrap(~apprehension_aor)

p1.1

p2 <- arrest_df %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  filter(yearmon >= as.yearmon("2025-01-01"),
         apprehension_aor == "Seattle Area of Responsibility") %>% 
  count(yearmon, third_country_deport) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_deport)) +
  geom_line()

p2

seattle_third_country_arrests <- arrest_df %>% 
  filter(apprehension_aor == "Seattle Area of Responsibility",
         unique_identifier %in% unlist(third_country_ids))

# Note possible multiple arrests per individual
tab1 <- seattle_third_country_arrests %>%
  filter(!is.na(departure_country)) %>% 
  count(citizenship_country, departure_country) %>% 
  arrange(desc(n))

knitr::kable(head(tab1, 10))

```

Here we check number of apparent third-country deportations based on characteristics of single dataset versus ids from all datasets; results appear to be same for Seattle AOR when querying same dataset:

```{r check_ids}

p2 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(yearmon = as.yearmon(departed_date)) %>% 
  count(yearmon, third_country_deport) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_deport)) +
  geom_line()

p2

p2.1 <- removals_df %>% 
  filter(docket_aor == "Seattle Area of Responsibility") %>% 
  mutate(yearmon = as.yearmon(departed_date),
         third_country_id = unique_identifier %in% third_country_ids) %>% 
  count(yearmon, third_country_id) %>% 
  ggplot(aes(x = yearmon, y = n, color = third_country_id)) +
  geom_line()

p2.1

```

```{r join_arrests_removals}

arrests_w_removal_data <- arrest_df %>% 
  filter(!is.na(departure_country))



```